---
title: "Cell-type perturbation in MASH vs healthy liver"
author: "Richard Ahn, PhD"
format:
  html:
    embed-resources: true
    toc: true
    toc-location: left
    code-fold: show
    theme: cosmo
    df-print: paged
execute:
  warning: false
  message: false
engine: knitr
---

```{r}
#| label: setup
#| code-fold: show

suppressPackageStartupMessages({
  library(Seurat)
  library(Matrix)
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(patchwork)
  library(speckle)     # propeller: donor-level differential abundance
  library(DESeq2)      # pseudobulk differential expression
  library(fgsea)       # GSEA on the ranked DE statistics
  library(msigdbr)     # MSigDB Hallmark gene sets
  library(knitr)
})

set.seed(1234)

# Format p-values for inline prose. signif() alone yields values like 1.3e-07,
# which the inline hook renders as broken superscript markup.
fmt_p <- function(x) {
  x <- unname(x[1])
  if (is.na(x)) return("NA")
  if (x < 1e-3) formatC(x, format = "e", digits = 1) else formatC(signif(x, 2), format = "g")
}

CACHE_DIR <- "cache"
dir.create(CACHE_DIR, showWarnings = FALSE, recursive = TRUE)
cache_path <- function(name) file.path(CACHE_DIR, name)

# Compute-or-load guard. Expensive steps that touch the full object write a
# small intermediate to cache/ and reuse it on later renders. Deleting cache/
# forces a full cold run, so the document stays reproducible from the raw data.
cached <- function(name, expr) {
  p <- cache_path(name)
  if (file.exists(p)) return(readRDS(p))
  x <- force(expr)
  saveRDS(x, p)
  x
}

# COL1A1 counts for every nucleus, needed in two separate sections. Wrapped in a
# function so neither section depends on the other having run first.
col1a1_counts <- function() cached("col1a1_distribution.rds", {
  library(BPCells)
  as.vector(as.matrix(open_matrix_dir("bpcells_counts")["COL1A1", ]))
})
```

## Introduction

The prompt is to take a MASH scRNA-seq or snRNA-seq dataset and highlight which cell types are
most perturbed and how relative to healthy liver. I think "most perturbed" carries two
distinct meanings:

1. **Differential abundance.** Which cell types change in proportion between MASH and
   control. A cell type can be strongly perturbed here while its surviving
   cells look transcriptionally ordinary.
2. **Within-cell-type differential expression.** How the cells of a given type are
   transcriptionally rewired in disease. A cell type can hold steady in
   proportion while its program shifts substantially.

These are measured separately below and only combined at the end, in a ranked
table that answers the prompt on both axes.

The cohort is staged rather than binary: healthy control, isolated steatosis
(MASL), early MASH and advanced MASH. The primary comparison is the clean
case-control contrast of control against MASH, with MASL held out so that
cases with steatosis but no steatohepatitis do not dilute it. MASL is then put
back to work in two secondary views: an ordered trend across all four stages,
and stage-adjacent contrasts that localize where along the progression each
cell type actually changes. A cell type that shifts only between early and
advanced MASH is telling a different story from one that shifts at the onset
of steatosis, and the two-group contrast alone cannot distinguish them.

I made the decision to run **analyses at the
donor level, not the cell level.** A dataset of a quarter of a million nuclei from 48 people
now carries 48 independent observations, not 250,000. Abundance is therefore tested on donor-level proportions, and
expression on donor-level pseudobulk.

I used Claude Code Opus 5 for this report.

## Executive summary

**Hepatocytes are the most perturbed cell type, and the only one perturbed on
both axes.** They lose more than half their share of nuclei across the
progression from control to MASH, and those nuclei carry more differentially expressed genes than
any other cell type by a wide margin. The program is coherent: p53 and growth
arrest up, differentiated metabolic function down, with bile acid, xenobiotic
and fatty acid metabolism all depleted. The compositional result independently
reproduces the source paper's headline.

**Endothelial cells do not change in abundance at all**, yet
carry the second-largest expression program in the dataset, and its content is
inflammatory activation. T cells expand several-fold in
number while changing their expression program comparatively little.

**Macrophages change identity rather than number,** switching from the resident
Kupffer program to a lipid-handling, scar-associated one. This is the most
mechanistically specific result here, and it reproduces the
macrophage arm of the fibrotic niche described in cirrhosis by Ramachandran et
al., in a cohort that has not reached cirrhosis.

**Two results required correcting the deposited annotation rather than trusting
it.** A subcluster labeled as inactivated stellate cells is vascular mural
tissue, which inverts the apparent direction of the stellate contractile
program when the label is aggregated. And stellate abundance appears unchanged
in the two-group contrast because it is not monotonic: the compartment expands
only at advanced fibrosis, which a control-versus-MASH comparison cancels out.

**What this dataset cannot answer.** It does not contain collagen-producing
myofibroblasts, and two of the three scar-associated populations from the
reference cirrhosis atlas are absent while the one not embedded in scar
replicates cleanly. I think the explanation is that needle-core biopsy
tissue does not recover cells resident in dense matrix. The consequence is nothing here bears on myofibroblast biology in either direction, and the flat collagen result is a sampling limit rather than evidence against
fibrogenesis.

**Implications for therapeutics are limited to target engageability.** The
receptors for thyromimetic, FXR and PPAR-alpha agonism are well expressed in the
cell types those drugs act on, and the programs they regulate are among the
things disease has disturbed. GLP-1 and GIP receptors are not expressed in this
tissue, so a liver atlas is the wrong instrument for those agents. However, the glucagon
arm of a dual agonist is the exception, with a direct and measurable hepatocyte
target. Anti-fibrotic targets cannot be assessed, for the sampling reason
above. None of this speaks to potential efficacy.

## Data provenance

Single-nucleus RNA-seq of 48 human liver biopsies spanning the MASLD spectrum,
from Hong et al., *Nature Genetics* 2025. Raw data are deposited under GEO
GSE289173 (BioProject PRJNA1221860); the processed Seurat object analyzed here
is distributed via Zenodo (10.5281/zenodo.14586466, CC-BY 4.0).

### Why this dataset

Several candidate datasets were considered before settling on this one (see table below). I chose the Hong et al. dataset because it is snRNA-seq, includes MASH and control samples from human liver biopsies, has a sufficient number of donors and nuclei, and, decisively, because its processed data was verified as downloadable rather than merely deposited.

::: {.column-page}

| #   | Dataset (accession)                                               | Human         | Liver biopsy | Indication            | sc / sn       | Atlas         | Download verified?                                                           |
| --- | ----------------------------------------------------------------- | ------------- | ------------ | --------------------- | ------------- | ------------- | ---------------------------------------------------------------------------- |
| 1   | sc-eQTL MASH, Hong 2025 (GSE289173 / Zenodo 14586466)             | Yes           | Yes          | MASH + control        | sn            | No (cohort)   | **Yes — processed Seurat v4 on Zenodo, CC-BY, 6.6 GB, opened and confirmed** |
| 2   | EphB2 NASH transition, Xiao 2023 (GSE189600)                      | Yes (+ mouse) | Yes          | NAFL→NASH             | sn (+ snATAC) | No            | Strongly — independently re-downloaded from GEO by ≥3 later papers           |
| 3   | Spatial multi-omics MASLD atlas, NatGen 2025 (s41588-025-02407-8) | Yes           | Yes          | control / MASL / MASH | both          | Yes           | Not yet — accession not confirmed                                            |
| 4   | Human HSC multi-omics, J Hepatol 2024 (S0168-8278(24)02667-9)     | Yes           | Yes          | normal / MASL / MASH  | sn (+ snATAC) | No (HSC-only) | Not yet — accession not confirmed                                            |
| 5   | Human snRNA-seq MASH (GSE212837)                                  | Yes           | Yes          | MASH                  | sn            | No            | Deposited + reused; file listing not opened                                  |
| 6   | Human scRNA-seq MASH (GSE189175)                                  | Yes           | Yes          | MASH                  | sc            | No            | Deposited + reused; file listing not opened                                  |
| —   | Copenhagen NASH, Fred 2022                                        | Yes           | Yes          | NASH                  | sc            | No            | **Excluded — no raw data released, no healthy controls**                     |

:::

The Hong et al. study is primarily a single-cell eQTL analysis, and its headline result is
a hepatocyte maladaptation signal at *EFHD1*/*FOXO1*. The cell-type
perturbation question asked here is described in that paper but not developed
in depth, and its differential expression was run cell-level with Seurat
`FindMarkers`. The contribution of this report is to answer the perturbation
question directly, with donor-level replication on both axes.

The downloadable object was built with Cellranger, Cellbender for ambient RNA,
DoubletFinder for doublets, Liger for integration and Seurat for clustering.
Liger was not an arbitrary choice: the authors benchmarked it against FastMNN,
Harmony, CCA and RPCA on kBET acceptance rate and selected it on that basis.
Normalization was done with `scran` and `batchelor` rather than Seurat's
log-normalization.
This analysis reuses that object and its integration as-is, and
validates the annotations rather than re-deriving them. 

### Cohort, contrasts and marker panels

The Zenodo deposit ships tidied metadata: seven columns, with disease stage
already in current MASLD/MASH nomenclature rather than the pre-2023 NAFLD terms
used in the authors' analysis scripts. No remapping is needed.

Two annotation columns are supplied. `Cell_type_broad` gives 14 labels;
`Cell_type_detailed` gives 19, and nests cleanly inside the broad column. Only
two lineages subdivide, and both subdivide along exactly the axis MASH acts on:
hepatocytes split by zonation, stellate cells by activation state. That is used
below, after the primary analysis, to ask where within a lineage the
perturbation sits.

```{r}
#| label: constants
#| code-fold: show

# Metadata columns, confirmed against the object rather than assumed.
COL_DONOR    <- "DonorID"
COL_STAGE    <- "Disease_group"
COL_CT_BROAD <- "Cell_type_broad"
COL_CT_FINE  <- "Cell_type_detailed"

# Disease stage, in the order of progression.
STAGE_LEVELS <- c("control", "MASL", "eMASH", "aMASH")
STAGE_LABELS <- c(control = "No-MASLD", MASL = "MASL",
                  eMASH = "eMASH", aMASH = "aMASH")

# Primary contrast is two-group. MASL is the intermediate stage and is held out
# of it, so the headline case-control comparison is not diluted by cases that
# have steatosis without steatohepatitis. MASL is not discarded: it carries the
# ordered trend analysis and the stage-adjacent contrasts defined below.
CONDITION_MAP <- c(control = "Control", MASL = NA,
                   eMASH = "MASH", aMASH = "MASH")
CONDITION_LEVELS <- c("Control", "MASH")

# Secondary stage-wise contrasts, run after the primary. Each is a step along
# the progression, which localizes where in the disease course a cell type
# changes rather than only whether it differs at the endpoint.
STAGE_CONTRASTS <- list(
  c("control", "MASL"),   # onset of steatosis
  c("MASL",    "eMASH"),  # steatosis to steatohepatitis
  c("control", "eMASH"),  # control to early MASH
  c("eMASH",   "aMASH"),  # early to advanced fibrosis
  c("control", "aMASH")   # full span, largest expected effect
)

# Thresholds for calling a gene differentially expressed. Significance alone
# admits very small effects when donor numbers are adequate, so an effect-size
# floor is applied alongside it. Both are stated wherever counts are reported.
FDR_CUTOFF <- 0.05
LFC_CUTOFF <- 0.25

# Applied to every DESeq2 result table in this report.
is_de <- function(r) {
  !is.na(r$padj) & r$padj < FDR_CUTOFF & abs(r$log2FoldChange) >= LFC_CUTOFF
}
DE_RULE <- paste0("FDR < ", FDR_CUTOFF, " and |log2FC| >= ", LFC_CUTOFF)

# Not cell types. Per the source paper's supplementary methods, "Unidentified"
# marks cells whose substate assignment disagreed between the joint and the
# disease-stratified subclustering. Excluded from every analysis below.
DROP_LABELS <- c("Suspected_doublet", "Unidentified")

# Canonical human liver markers, used to verify the shipped annotations before
# any claim is built on them.
LIVER_MARKERS <- list(
  Hepatocyte    = c("ALB", "APOA1", "TTR", "CYP2E1", "HNF4A", "ASGR1"),
  Cholangiocyte = c("KRT19", "KRT7", "EPCAM", "SOX9", "CFTR"),
  Stellate      = c("COL1A1", "COL3A1", "DCN", "PDGFRB", "RGS5"),
  Endothelial   = c("PECAM1", "VWF", "CLEC4G", "STAB2", "LYVE1"),
  Kupffer_Mac   = c("CD68", "MARCO", "VSIG4", "C1QA", "CD5L"),
  Scar_Mac      = c("TREM2", "CD9", "GPNMB", "SPP1", "LYZ"),
  T_cell        = c("CD3D", "CD3E", "CD8A", "IL7R"),
  NK            = c("NKG7", "GNLY", "KLRD1"),
  B_Plasma      = c("CD79A", "MS4A1", "MZB1", "IGHG1")
)

# programs the differential expression results are checked against. These are
# recovered from the data, not asserted; they exist here to make the check
# explicit rather than post hoc.
MASH_PROGRAMS <- list(
  HSC_activation      = c("COL1A1", "COL3A1", "ACTA2", "TIMP1", "TAGLN"),
  Scar_macrophage     = c("TREM2", "CD9", "GPNMB", "SPP1"),
  Kupffer_homeostatic = c("MARCO", "VSIG4", "CD5L", "CLEC4F"),
  LSEC_identity       = c("CLEC4G", "STAB2", "LYVE1", "OIT3"),
  Ductular_reaction   = c("KRT19", "KRT7", "SOX9", "EPCAM")
)
```

## Loading the object

The deposited object is Seurat v4 and occupies roughly 14 GB once loaded, which
leaves no headroom on a 16 GB machine. That shapes the code below, especially the decision to run on-disk using BPCells.

Everything expensive runs once, writes a compact intermediate to `cache/`, and
is reused afterwards. Counts are additionally written to an on-disk BPCells
matrix, so per-cell-type aggregation streams from disk instead of holding a
24,615 x 249,233 sparse matrix in RAM. Deleting `cache/` and `bpcells_counts/`
forces a full recomputation from the deposited file.

```{r}
#| label: heavy-pass
#| code-fold: show

RDS <- "data/unzipped/Zenodo/processed_seurat_object/MASLD_snRNA_seq_seurat_v4.rds"

if (!file.exists(cache_path("meta.rds")) || !dir.exists("bpcells_counts")) {
  library(BPCells)
  mem.maxVSize(vsize = 64000)          # allow the load to spill past physical RAM
  obj <- readRDS(RDS)
  rna <- obj@assays$RNA

  saveRDS(obj@meta.data, cache_path("meta.rds"))
  saveRDS(list(dim = dim(obj), reductions = names(obj@reductions),
               genes = rownames(rna), seurat_ver = as.character(obj@version)),
          cache_path("obj_info.rds"))
  for (rd in names(obj@reductions)) {
    e <- obj@reductions[[rd]]@cell.embeddings
    saveRDS(e[, 1:min(3, ncol(e))], cache_path(paste0("emb_", rd, ".rds")))
  }

  cnt <- rna@counts                     # raw integer counts, required for pseudobulk
  rm(obj, rna); gc()
  BPCells::write_matrix_dir(cnt, dir = "bpcells_counts", compress = TRUE)
  rm(cnt); gc()
}

md   <- readRDS(cache_path("meta.rds"))
info <- readRDS(cache_path("obj_info.rds"))

md$stage     <- factor(md[[COL_STAGE]], levels = STAGE_LEVELS,
                       labels = STAGE_LABELS[STAGE_LEVELS])
md$condition <- factor(unname(CONDITION_MAP[as.character(md[[COL_STAGE]])]),
                       levels = CONDITION_LEVELS)

# Technical labels are not cell types and are excluded from every analysis.
n_dropped <- sum(md[[COL_CT_BROAD]] %in% DROP_LABELS)
md <- md[!md[[COL_CT_BROAD]] %in% DROP_LABELS, ]
```

The object holds `r format(info$dim[2], big.mark = ",")` nuclei and
`r format(info$dim[1], big.mark = ",")` genes, integrated with Liger
(reductions: `r paste(info$reductions, collapse = ", ")`). Dropping
`r paste(DROP_LABELS, collapse = " and ")` removes
`r format(n_dropped, big.mark = ",")` nuclei, leaving
`r format(nrow(md), big.mark = ",")` for analysis.

## Cohort

```{r}
#| label: cohort

donors <- unique(md[, c(COL_DONOR, "stage", "condition")])
names(donors)[1] <- "donor"

cohort <- as.data.frame(table(Stage = donors$stage), responseName = "Donors")
cohort$Nuclei <- as.integer(table(factor(md$stage, levels = levels(donors$stage))))
cohort$`Median nuclei/donor` <- round(tapply(
  as.integer(table(md[[COL_DONOR]]))[donors$donor], donors$stage, median))
cohort$`Primary contrast` <- ifelse(
  is.na(CONDITION_MAP[STAGE_LEVELS]), "held out",
  as.character(CONDITION_MAP[STAGE_LEVELS]))
kable(cohort, caption = "Cohort by disease stage.")
```

`r nrow(donors)` donors, distributed as
`r paste(sprintf("%s n=%d", cohort$Stage, cohort$Donors), collapse = ", ")`.
This reproduces the cohort reported in the source paper exactly.

Note that the stages are not balanced in nuclei: controls contribute
`r round(100 * sum(md$stage == "No-MASLD") / nrow(md))`% of nuclei but only
`r round(100 * sum(donors$stage == "No-MASLD") / nrow(donors))`% of donors.
Pooling nuclei across donors would therefore weight the control group nearly
twice as heavily as its donor count warrants, which is the first practical
reason the analyses below operate on per-donor quantities.

One further detail matters for anyone reusing this object: donor identifiers do
not encode disease stage. All 20 `ctrl_*` donors are controls, but three further
controls are named `FL*`. Condition must be taken from `r COL_STAGE`, never
parsed from the donor name.

```{r}
#| label: naming-check
kable(table(`Donor name` = ifelse(grepl("^ctrl", donors$donor), "ctrl_*", "FL*"),
            Stage = donors$stage),
      caption = "Donor naming does not track disease stage.")
```

## QC and study design checks

This is validation of a processed object, not re-derivation. The object has
already been through Cellbender ambient-RNA removal and DoubletFinder. The
question is whether anything about its quality differs systematically between
conditions in a way that could masquerade as biology.

```{r}
#| label: qc
#| fig-width: 10
#| fig-height: 3.4

qc <- tidyr::pivot_longer(
  md[, c("condition", "stage", "nCount_RNA", "nFeature_RNA", "percent.mt")],
  cols = c(nCount_RNA, nFeature_RNA, percent.mt),
  names_to = "metric", values_to = "value")

ggplot(subset(qc, !is.na(condition)), aes(condition, value, fill = condition)) +
  geom_violin(scale = "width", linewidth = 0.3) +
  geom_boxplot(width = 0.12, outlier.shape = NA, linewidth = 0.3, fill = "white") +
  facet_wrap(~ metric, scales = "free_y") +
  scale_y_continuous(trans = "log1p") +
  scale_fill_manual(values = c(Control = "#4C72B0", MASH = "#C44E52")) +
  labs(x = NULL, y = "value (log1p scale)") +
  theme_bw(base_size = 11) + theme(legend.position = "none")
```

For reference, the filters already applied by the source study, from its
supplementary methods: singlets were retained on gene and read count with a
per-donor threshold (count greater than 0.2 times that donor's median) and genes
expressed in fewer than 0.25% of cells in a donor were removed. At sample level,
donors with a median below 200 genes per cell were excluded, along with one
principal-component outlier and three samples failing genome-transcriptome
concordance. Ambient RNA was removed with Cellbender and doublets with
DoubletFinder before any of that. The checks below are therefore validation of
an already-filtered object, not a second round of filtering.

`percent.mt` is essentially zero across the dataset (median
`r round(median(md$percent.mt), 3)`, maximum
`r round(max(md$percent.mt), 1)`%). That is the expected behavior for
single-nucleus data, where mitochondrial transcripts are largely excluded with
the cytoplasm. Mitochondrial filtering is not a meaningful QC axis here, and
applying a conventional 5-10% threshold wouldn't really be meaningful.

Depth differs modestly between conditions, which is worth knowing but is handled
downstream: pseudobulk DE models library size explicitly through DESeq2's size
factors, and proportions are depth-invariant by construction.

### Confounding

The design can only adjust for variables that vary *within* condition. Two
checks decide the model formula.

```{r}
#| label: confounds

# Does any donor contribute more than one condition?
per_donor <- table(md[[COL_DONOR]], md[[COL_STAGE]])
multi <- sum(rowSums(per_donor > 0) > 1)

meta_cols <- colnames(readRDS(cache_path("meta.rds")))

data.frame(
  Check = c("Metadata columns deposited",
            "Donors contributing >1 disease stage",
            "Batch / chemistry / center variable present"),
  Result = c(paste(meta_cols, collapse = ", "),
             as.character(multi),
             "none")
) |> kable(caption = "Design checks determining the DE formula.")
```

Every donor contributes exactly one disease stage, so **donor is nested within
condition**. It therefore cannot enter the design: a donor term would absorb the
between-donor variation that *is* the condition effect. This is a correctness
constraint rather than a modeling preference.

The deposited metadata carries no batch, chemistry, sequencing-run or center
variable. There is consequently nothing to adjust for, and the design stays
`~ condition`. This is a single-center Korean cohort, so the risk of hidden
batch-condition confounding is lower than in a multi-dataset atlas.

## Verifying annotations

Every claim below rests on the deposited labels, so they are checked against
canonical liver markers before use. Expression is summarized per cell type from
the on-disk counts: library-size normalized to 10,000, `log1p` transformed.

```{r}
#| label: markers
#| fig-width: 10
#| fig-height: 5.5

mk <- cached("markers_broad.rds", {
  library(BPCells)
  mat  <- open_matrix_dir("bpcells_counts")
  meta <- readRDS(cache_path("meta.rds"))
  keep <- !meta[[COL_CT_BROAD]] %in% DROP_LABELS
  ct   <- factor(as.character(meta[[COL_CT_BROAD]])[keep])
  sub  <- mat[, which(keep)]
  normed <- log1p(multiply_cols(sub, 1e4 / colSums(sub)))
  genes  <- intersect(unique(unlist(LIVER_MARKERS)), rownames(mat))
  avg <- pseudobulk_matrix(normed[genes, ], cell_groups = ct, method = "mean")
  nz  <- pseudobulk_matrix(sub[genes, ],    cell_groups = ct, method = "nonzeros")
  pct <- sweep(nz, 2, as.integer(table(ct)), "/") * 100
  flat <- utils::stack(LIVER_MARKERS)
  data.frame(gene = rep(rownames(avg), ncol(avg)),
             cell_type = rep(colnames(avg), each = nrow(avg)),
             avg_exp = as.vector(avg), pct_exp = as.vector(pct),
             panel = as.character(flat$ind[match(rep(rownames(avg), ncol(avg)),
                                                 flat$values)]))
})

# Scale each gene across cell types so the dotplot shows specificity, not
# absolute abundance: a gene high everywhere is not evidence for any label.
mk <- mk |>
  group_by(gene) |>
  mutate(z = as.numeric(scale(avg_exp))) |>
  ungroup()
mk$gene      <- factor(mk$gene, levels = intersect(unlist(LIVER_MARKERS), mk$gene))
mk$cell_type <- factor(mk$cell_type, levels = rev(sort(unique(mk$cell_type))))

ggplot(mk, aes(gene, cell_type)) +
  geom_point(aes(size = pct_exp, colour = z)) +
  scale_colour_gradient2(low = "#3B4CC0", mid = "grey92", high = "#B40426",
                         midpoint = 0, name = "scaled\nexpression") +
  scale_size_continuous(range = c(0.2, 5), name = "% expressing") +
  labs(x = NULL, y = NULL) +
  theme_bw(base_size = 10) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
        panel.grid.major = element_line(linewidth = 0.2))
```

Each label expresses its canonical markers: hepatocytes *ALB*, *ASGR1*,
*CYP2E1*, *HNF4A*; endothelial cells the sinusoidal program *STAB2*,
*CLEC4G*, *LYVE1*; stellate cells *DCN*, *COL3A1*, *PDGFRB*; macrophages the
Kupffer set *MARCO*, *VSIG4*, *CD5L*; and the lymphoid labels their respective
*CD3*, *KLRD1*/*GNLY* and *MS4A1*/*CD79A* markers.

Two labels warrant an explicit note. Monocytes score highest for *PECAM1*, which
initially looks like endothelial contamination. It is not: *PECAM1* is CD31,
genuinely expressed on monocytes, and the population carries monocyte-restricted
*FCN1* and *CSF1R* while lacking the endothelial-restricted *VWF* and *EGFL7*.
Dendritic cells likewise carry *FLT3*, *IRF8* and *CD1C*. Both labels are sound.
They are nonetheless the two smallest populations
(`r format(sum(md[[COL_CT_BROAD]] == "Monocyte"), big.mark = ",")` and
`r format(sum(md[[COL_CT_BROAD]] == "DC"), big.mark = ",")` nuclei), which
limits them later on power grounds rather than on annotation quality.

`Other_T` and `Stellate_cell_misc` are heterogeneous residual labels. They are
retained so that proportions sum over the whole tissue, but no biological
conclusion is drawn from their expression results.

### The detailed labels

The deposit also ships a finer annotation, which subdivides only two lineages:
hepatocytes by zonation and stellate cells by state. The same marker panel is
applied to it.

```{r}
#| label: markers-detailed
#| fig-width: 10
#| fig-height: 6

mk_fine <- cached("markers_detailed.rds", {
  library(BPCells)
  mat  <- open_matrix_dir("bpcells_counts")
  meta <- readRDS(cache_path("meta.rds"))
  keep <- !meta[[COL_CT_FINE]] %in% DROP_LABELS
  ctf  <- factor(as.character(meta[[COL_CT_FINE]])[keep])
  sub  <- mat[, which(keep)]
  normed <- log1p(multiply_cols(sub, 1e4 / colSums(sub)))
  genes  <- intersect(unique(unlist(LIVER_MARKERS)), rownames(mat))
  avg <- pseudobulk_matrix(normed[genes, ], cell_groups = ctf, method = "mean")
  nz  <- pseudobulk_matrix(sub[genes, ],    cell_groups = ctf, method = "nonzeros")
  pct <- sweep(nz, 2, as.integer(table(ctf)), "/") * 100
  data.frame(gene = rep(rownames(avg), ncol(avg)),
             cell_type = rep(colnames(avg), each = nrow(avg)),
             avg_exp = as.vector(avg), pct_exp = as.vector(pct))
})

mk_fine <- mk_fine |> group_by(gene) |>
  mutate(z = as.numeric(scale(avg_exp))) |> ungroup()
mk_fine$gene <- factor(mk_fine$gene, levels = intersect(unlist(LIVER_MARKERS), mk_fine$gene))
mk_fine$cell_type <- factor(mk_fine$cell_type, levels = rev(sort(unique(mk_fine$cell_type))))

ggplot(mk_fine, aes(gene, cell_type)) +
  geom_point(aes(size = pct_exp, colour = z)) +
  scale_colour_gradient2(low = "#3B4CC0", mid = "grey92", high = "#B40426",
                         midpoint = 0, name = "scaled\nexpression") +
  scale_size_continuous(range = c(0.2, 5), name = "% expressing") +
  labs(x = NULL, y = NULL) +
  theme_bw(base_size = 10) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
        panel.grid.major = element_line(linewidth = 0.2))
```

The three hepatocyte zones share the hepatocyte marker profile, as they should:
zonation is a gradient within a lineage, not a change of identity. The stellate
substates mostly share the stellate profile.

One does not, and it matters for how the stellate results are read.

```{r}
#| label: mural-check

sub_expr <- cached("stellate_subtype_expr.rds", {
  library(BPCells)
  mat  <- open_matrix_dir("bpcells_counts")
  meta <- readRDS(cache_path("meta.rds"))
  sel  <- grepl("^Stellate", meta[[COL_CT_FINE]])
  ctf  <- factor(as.character(meta[[COL_CT_FINE]])[sel])
  sb   <- mat[, which(sel)]
  nm   <- log1p(multiply_cols(sb, 1e4 / colSums(sb)))
  g <- c("DCN", "RBP1", "LRAT", "CYGB", "HGF", "COL1A1", "COL1A2",
         "PDGFRB", "TAGLN", "ACTA2", "PPARG",
         "MYH11", "MYLK", "MYL9", "NOTCH3", "RGS5")
  g <- g[g %in% rownames(mat)]
  t(pseudobulk_matrix(nm[g, ], cell_groups = ctf, method = "mean"))
})
rownames(sub_expr) <- sub("Stellate_cell_", "", rownames(sub_expr))

kable(round(sub_expr[, c("LRAT", "DCN", "RBP1", "ACTA2", "TAGLN",
                         "MYH11", "NOTCH3", "RGS5")], 2),
      caption = "Stellate subclusters. Left: hepatic stellate identity. Right: smooth-muscle and pericyte genes.")
```

The subcluster labeled `Stellate_cell_inactivated` carries
`r round(sub_expr["inactivated", "MYH11"] / max(sub_expr[rownames(sub_expr) != "inactivated", "MYH11"]))`-fold
more *MYH11* than any other stellate subcluster, plus the only appreciable
*NOTCH3*, while holding the **lowest** *LRAT* and *DCN* of the four. *MYH11* is
smooth-muscle myosin heavy chain and is not expressed by hepatic stellate cells
in any activation state, and a genuinely inactivated stellate cell that has
reverted after activation retains its lipid-storing identity markers. This
population has neither.

It is therefore better described as vascular smooth muscle or pericyte than as a
stellate cell of any state, and it is treated that way in the detailed
differential expression below, **where it is relabeled `Mural_VSMC`**. The
consequences of the mislabel for the aggregate stellate results are worked
through in the supplementary material.

### Testing for myofibroblasts

**No myofibroblast population is recovered, which bounds every fibrosis claim.**
*TREM2* is effectively undetected here despite being present in the object, and
the fibrillar collagens do not move significantly in stellate cells. It would be
easy to attribute both to single-nucleus chemistry. For *TREM2* that is
reasonable. For the collagens it is not, and the distinction matters enough to
test rather than assume.

```{r}
#| label: snrna-sensitivity

detect <- cached("snrna_detection.rds", {
  library(BPCells)
  mt   <- open_matrix_dir("bpcells_counts")
  meta <- readRDS(cache_path("meta.rds"))
  sel  <- meta[[COL_CT_BROAD]] == "Stellate_cell"
  sb   <- mt[, which(sel)]
  one  <- rep("all", ncol(sb))
  nz   <- as.vector(pseudobulk_matrix(sb, cell_groups = one, method = "nonzeros"))
  su   <- as.vector(pseudobulk_matrix(sb, cell_groups = one, method = "sum"))
  tot  <- colSums(sb)
  list(tab = data.frame(gene = rownames(sb), pct = nz / ncol(sb) * 100,
                        cpm = su / sum(su) * 1e6),
       n_cells = ncol(sb), median_umi = median(tot))
})

col_dist <- col1a1_counts()

d <- detect$tab; d$rank <- rank(-d$cpm)
g <- c("COL3A1", "COL1A2", "COL1A1", "COL4A1", "FN1")
i <- match(g, d$gene)

# Detection expected from sequencing depth alone, if counts were Poisson.
expected <- (1 - exp(-d$cpm[i] * detect$median_umi / 1e6)) * 100

kable(data.frame(
  Gene = g,
  CPM = round(d$cpm[i]),
  `Rank of 24,615` = round(d$rank[i]),
  `% nuclei detected` = round(d$pct[i], 1),
  `% expected from depth` = round(expected, 1),
  check.names = FALSE), row.names = FALSE,
  caption = "Observed detection in stellate nuclei against what sequencing depth alone predicts.")
```

Observed detection tracks the depth expectation closely for every collagen
tested, so there is no evidence of selective loss of this transcript class.
*COL3A1* is the
`r round(d$rank[d$gene == "COL3A1"])`th most abundant gene in stellate nuclei and
is detected in `r round(d$pct[d$gene == "COL3A1"], 1)`% of them. Collagen
transcripts are clearly capturable in this preparation.

**The problem is that the cells that should express *COL1A1* highly are not
present!**

```{r}
#| label: col1a1-absence
kable(data.frame(
  `COL1A1 counts per nucleus` = c("0", "1-2", "3-4", "5-9", "10 or more"),
  `Nuclei` = c(sum(col_dist == 0), sum(col_dist %in% 1:2), sum(col_dist %in% 3:4),
               sum(col_dist %in% 5:9), sum(col_dist >= 10)),
  check.names = FALSE), row.names = FALSE,
  caption = "COL1A1 across all nuclei in the dataset.")
```

The highest *COL1A1* count in any nucleus in the entire dataset is
`r max(col_dist)`, and only `r sum(col_dist >= 10)` nuclei carry ten or more. In
a collagen-producing myofibroblast, *COL1A1* is among the most abundant
transcripts in the library by a wide margin. A dataset-wide maximum of
`r max(col_dist)` counts is not a detection failure, it is the absence of the
cell type.

Two things most plausibly explain that absence, and neither is assay chemistry.
Myofibroblasts in fibrotic tissue are embedded in dense cross-linked matrix and
are recovered poorly by dissociation, a bias that is worse the less tissue there
is to work with and these are from needle core biopsy samples. And the cohort tops out at
F3-F4 fibrosis in ten donors, which is not end-stage disease. Consistent with
both, the stellate compartment is only
`r round(100 * detect$n_cells / nrow(md), 1)`%
of the nuclei analyzed, and stellate nuclei are among the most shallowly
sequenced in the dataset at a median of
`r format(round(detect$median_umi), big.mark = ",")` UMIs.

The consequence for reading this report is specific. Abundance and
pathway-level results do not depend on recovering myofibroblasts. Statements
about stellate cells describe the hepatic stellate populations that were
recovered, which are largely quiescent and activated states rather than
matrix-producing myofibroblasts. Nothing here should be taken as evidence about
myofibroblast biology in MASH, in either direction, and the flat collagen result
in particular is not evidence against fibrogenesis. A dataset built on larger
tissue pieces, precision-cut slices (as I've done in the past) or resections rather than cores, would be
the appropriate design for that question.


# Axis 1: which cell types change in abundance

## Composition across the disease spectrum

```{r}
#| label: composition
#| fig-width: 11
#| fig-height: 4.6
#| layout-nrow: 2

props <- cached("donor_props.rds", {
  tab <- table(md[[COL_DONOR]], md[[COL_CT_BROAD]])
  list(counts = tab, props = sweep(tab, 1, rowSums(tab), "/"))
})$props

dstage <- donors$stage[match(rownames(props), donors$donor)]

long <- data.frame(
  donor = rep(rownames(props), ncol(props)),
  cell_type = rep(colnames(props), each = nrow(props)),
  prop = as.vector(props),
  stage = rep(dstage, ncol(props)))

# Order donors within stage by hepatocyte fraction, so the gradient is visible.
hep <- props[, "Hepatocyte"]
long$donor <- factor(long$donor, levels = rownames(props)[order(dstage, -hep)])

pal <- setNames(
  c("#E8743B", "#19A979", "#945ECF", "#13A4B4", "#525DF4", "#BF399E",
    "#6C8893", "#EE6868", "#2F6497", "#B4B9C4", "#8A5B2E", "#C9C300"),
  sort(unique(long$cell_type)))

# Aggregated view: mean of the per-donor proportions within each stage, so every
# donor contributes equally regardless of how many nuclei it yielded.
agg <- aggregate(prop ~ cell_type + stage, long, mean)

p0 <- ggplot(agg, aes(stage, prop, fill = cell_type)) +
  geom_col(width = 0.75) +
  scale_fill_manual(values = pal, name = NULL) +
  scale_y_continuous(expand = c(0, 0), labels = scales::percent) +
  labs(x = NULL, y = "mean proportion of nuclei") +
  theme_bw(base_size = 10) +
  theme(panel.grid = element_blank(), legend.key.size = unit(9, "pt"))
p0

p1 <- ggplot(long, aes(donor, prop, fill = cell_type)) +
  geom_col(width = 1) +
  facet_grid(~ stage, scales = "free_x", space = "free_x") +
  scale_fill_manual(values = pal, name = NULL) +
  scale_y_continuous(expand = c(0, 0), labels = scales::percent) +
  labs(x = "donor", y = "proportion of nuclei") +
  theme_bw(base_size = 10) +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        panel.grid = element_blank(), legend.key.size = unit(9, "pt"))
p1
```

The first panel aggregates within stage (and mirrors Figure 1b of Hong et al.), averaging the per-donor proportions so
that each donor counts once regardless of how many nuclei it contributed. The
second shows every donor separately. The hepatocyte compartment visibly contracts
across the progression while lymphoid and myeloid compartments fill the space, and the
effect is consistent enough to be apparent donor by donor rather than only on
group means.

```{r}
#| label: stage-means

stage_mean <- sapply(levels(donors$stage), function(s)
  colMeans(props[dstage == s, , drop = FALSE]) * 100)
stage_mean <- stage_mean[order(-stage_mean[, "No-MASLD"]), ]
kable(round(stage_mean, 2),
      caption = "Mean per-donor cell-type proportion (%) by disease stage.")
```

Hepatocytes fall from `r round(stage_mean["Hepatocyte", "No-MASLD"], 1)`% of
nuclei in controls to `r round(stage_mean["Hepatocyte", "aMASH"], 1)`% in
advanced MASH. The source paper reports approximately 48% to 21% for the same
comparison, so this recovers their headline compositional result independently.

## Differential abundance testing

`propeller` operates on per-donor proportions: logit-transformed, then
moderated t-tests with empirical Bayes shrinkage across cell types. The donor is
the unit throughout, so 48 donors give 48 observations, not 249,233.

```{r}
#| label: propeller-primary

library(speckle)
sub_md <- md[!is.na(md$condition), ]

prop_res <- cached("propeller_primary.rds", propeller(
  clusters = sub_md[[COL_CT_BROAD]],
  sample   = sub_md[[COL_DONOR]],
  group    = sub_md$condition))

prop_tab <- data.frame(
  `Cell type`  = as.character(prop_res$BaselineProp.clusters),
  `Control %`  = round(prop_res$PropMean.Control * 100, 2),
  `MASH %`     = round(prop_res$PropMean.MASH * 100, 2),
  `Fold change`= round(prop_res$PropMean.MASH / prop_res$PropMean.Control, 2),
  FDR          = signif(prop_res$FDR, 3),
  check.names  = FALSE)
prop_tab <- prop_tab[order(-prop_tab$`Fold change`), ]

# Significant cell types are marked in the table rather than left to the reader
# to derive from the FDR column.
prop_show <- prop_tab
sig_ct <- prop_show$FDR < FDR_CUTOFF
prop_show$`Cell type` <- ifelse(sig_ct, paste0("**", prop_show$`Cell type`, "**"),
                                prop_show$`Cell type`)
kable(prop_show, row.names = FALSE,
      caption = paste0("propeller, Control (n=23) vs MASH (n=19), ranked by fold change. ",
                       "Fold change is MASH relative to Control. Cell types in bold are ",
                       "significant at Benjamini-Hochberg FDR < ", FDR_CUTOFF, " (",
                       sum(sig_ct), " of ", nrow(prop_show), ")."))
```

```{r}
#| label: propeller-plot
#| fig-width: 10
#| fig-height: 4

# Every cell type significant in the primary contrast, ordered by fold change,
# so the panel and the table agree.
show <- prop_tab$`Cell type`[prop_tab$FDR < FDR_CUTOFF]
pl <- subset(long, cell_type %in% show & !is.na(stage))
pl$cell_type <- factor(pl$cell_type, levels = show)

ggplot(pl, aes(stage, prop, fill = stage)) +
  geom_boxplot(outlier.shape = NA, linewidth = 0.3) +
  geom_jitter(width = 0.15, size = 0.7, alpha = 0.6) +
  facet_wrap(~ cell_type, scales = "free_y", nrow = 2) +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = c("#4C72B0", "#DD8452", "#C44E52", "#8172B3")) +
  labs(x = NULL, y = "proportion of nuclei") +
  theme_bw(base_size = 10) +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 30, hjust = 1))
```

`r sum(prop_tab$FDR < FDR_CUTOFF)` of `r nrow(prop_tab)` cell types change significantly in abundance. The pattern is a
single coherent event: hepatocytes are lost
(`r round(prop_tab$"Fold change"[prop_tab$"Cell type" == "Hepatocyte"], 2)`-fold,
FDR `r fmt_p(prop_tab$FDR[prop_tab$"Cell type" == "Hepatocyte"])`) and every
significant immune population expands, led by CD4 T cells
(`r round(prop_tab$"Fold change"[prop_tab$"Cell type" == "CD4_T"], 1)`-fold) and
CD8 T cells
(`r round(prop_tab$"Fold change"[prop_tab$"Cell type" == "CD8_T"], 1)`-fold).

Because proportions are compositional, these are not independent measurements.
A large loss of hepatocytes mechanically raises every other fraction. The
immune expansion should therefore be read as relative enrichment, not as proof
of absolute immune infiltration, and that caveat applies to any proportion-based
analysis of this kind.

### Abundance across individual disease stages

The two-group contrast pools early and advanced MASH. Testing the stages against
each other separately shows where along the progression each shift occurs.

```{r}
#| label: stage-tables

stagewise <- cached("propeller_stagewise.rds", {
  out <- list()
  for (cc in STAGE_CONTRASTS) {
    sel <- md[[COL_STAGE]] %in% cc
    out[[paste(cc, collapse = " vs ")]] <- propeller(
      clusters = md[[COL_CT_BROAD]][sel],
      sample   = md[[COL_DONOR]][sel],
      group    = factor(as.character(md[[COL_STAGE]])[sel], levels = cc))
  }
  out
})

stage_table <- function(key) {
  r <- stagewise[[key]]
  cts <- as.character(r$BaselineProp.clusters)
  # propeller reports the ratio of the first group to the second; invert so the
  # column reads as change in the later stage relative to the earlier one.
  fc <- round(1 / r$PropRatio, 2)
  d <- data.frame(`Cell type` = ifelse(r$FDR < FDR_CUTOFF, paste0("**", cts, "**"), cts),
                  `Fold change` = fc, FDR = signif(r$FDR, 3), check.names = FALSE)
  d <- d[order(-d$`Fold change`), ]
  kable(d, row.names = FALSE,
        caption = paste0(key, ". Fold change is the later stage relative to the earlier. ",
                         "Bold: FDR < ", FDR_CUTOFF, " (", sum(r$FDR < FDR_CUTOFF),
                         " of ", nrow(r), ")."))
}
```

```{r}
#| label: stage-table-1
stage_table("control vs eMASH")
```

```{r}
#| label: stage-table-2
stage_table("control vs aMASH")
```

```{r}
#| label: stage-table-3
stage_table("eMASH vs aMASH")
```

Control against early MASH and control against advanced MASH both recover the
hepatocyte loss and lymphoid expansion seen in the pooled contrast. The
early-versus-advanced comparison is informative but nothing is signficant after FDR adjustment.

## Stage-resolved abundance: where the two-group test misleads

Stellate cells are the central fibrosis cell type, yet they are flat in the
two-group contrast
(FDR `r fmt_p(prop_tab$FDR[prop_tab$"Cell type" == "Stellate_cell"])`).
Splitting MASH back into its two stages explains why.

```{r}
#| label: stagewise

nonmono <- stage_mean[c("Stellate_cell", "Cholangiocyte"), ]
kable(round(nonmono, 2),
      caption = "Two cell types whose abundance is not monotonic across the progression (%).")
```

Both dip through early disease and then expand at advanced fibrosis. Averaging
eMASH and aMASH into a single MASH arm cancels the change, which is exactly why
the two-group test returns nothing for stellate cells.

```{r}
#| label: stagewise-stellate

ea <- stagewise[["eMASH vs aMASH"]]
kable(data.frame(
  `Cell type` = as.character(ea$BaselineProp.clusters),
  `Fold change (aMASH/eMASH)` = round(1 / ea$PropRatio, 2),
  `P` = signif(ea$P.Value, 3),
  FDR = signif(ea$FDR, 3),
  check.names = FALSE)[1:5, ], row.names = FALSE,
  caption = "Top five cell types, early vs advanced MASH (9 vs 10 donors).")
```

Stellate cells are the strongest signal in this contrast, expanding
`r round(1 / ea$PropRatio[ea$BaselineProp.clusters == "Stellate_cell"], 1)`-fold
from early to advanced MASH
(p `r fmt_p(ea$P.Value[ea$BaselineProp.clusters == "Stellate_cell"])`).

Again, this does not survive multiple-testing correction
(FDR `r fmt_p(ea$FDR[ea$BaselineProp.clusters == "Stellate_cell"])`), and the
comparison rests on 9 versus 10 donors. I think what's going on here is a suggestive stage-specific pattern that is consistent with fibrosis
biology and with the significant expression changes reported below, though not an
established differential abundance result. It is reported because a two-group-only analysis
would have recorded stellate cells as unchanged which isn't entirely true.

# Axis 2: how each cell type is transcriptionally perturbed

## Building donor-level pseudobulk

Counts are summed within each (cell type, donor) pair. This is the step that
makes the donor the replication unit: one pseudo-sample per donor per cell type,
carrying that donor's condition label exactly once.

```{r}
#| label: pseudobulk

MIN_CELLS  <- 20   # a pseudo-sample must aggregate at least this many nuclei
MIN_DONORS <- 10   # a cell type needs this many donors in each group to be tested

pb <- cached("pseudobulk_broad.rds", {
  library(BPCells)
  mat  <- open_matrix_dir("bpcells_counts")
  meta <- readRDS(cache_path("meta.rds"))
  keep <- !meta[[COL_CT_BROAD]] %in% DROP_LABELS
  grp  <- paste(as.character(meta[[COL_CT_BROAD]])[keep],
                as.character(meta[[COL_DONOR]])[keep], sep = "__")
  pseudobulk_matrix(mat[, which(keep)], cell_groups = grp, method = "sum")
})

pbmeta <- cached("pseudobulk_meta_broad.rds", {
  key <- do.call(rbind, strsplit(colnames(pb), "__", fixed = TRUE))
  m <- data.frame(pseudosample = colnames(pb), cell_type = key[, 1],
                  donor = key[, 2], stringsAsFactors = FALSE)
  m$n_cells <- as.integer(table(paste(md[[COL_CT_BROAD]], md[[COL_DONOR]],
                                      sep = "__"))[m$pseudosample])
  m$stage     <- donors$stage[match(m$donor, donors$donor)]
  m$condition <- donors$condition[match(m$donor, donors$donor)]
  m
})
pbmeta <- pbmeta[match(colnames(pb), pbmeta$pseudosample), ]

cov <- with(subset(pbmeta, n_cells >= MIN_CELLS & !is.na(condition)),
            table(cell_type, condition))
kable(cov, caption = paste0("Donors passing the ", MIN_CELLS,
      "-nucleus threshold, per cell type. Cell types below ", MIN_DONORS,
      " donors in either group are not tested."))
```

Monocytes and dendritic cells fall out here. They are genuine populations, as
the marker check established, but at a median of
`r round(median(subset(pbmeta, cell_type == "Monocyte")$n_cells))` and
`r round(median(subset(pbmeta, cell_type == "DC")$n_cells))` nuclei per donor
their pseudobulk profiles are dominated by sampling noise. They remain in the
abundance analysis, where donor proportions are still well estimated.

## Differential expression

DESeq2 per cell type, design `~ condition`.

A gene is called differentially expressed on two criteria together:
Benjamini-Hochberg FDR below `r FDR_CUTOFF`, and an absolute log2 fold change of
at least `r LFC_CUTOFF`. Significance alone is a statement about confidence, not
about size, and with 23 and 19 donors it admits changes too small to be
interesting. The effect-size floor is applied as a filter on the results rather
than as a null hypothesis in the model, so the reported p-values remain tests
against no change. Enrichment analyses later in the report are unaffected as they
rank all genes by the test statistic and never threshold.

Donor is deliberately absent from the design. Each donor contributes exactly one
condition, so donor is nested within condition; including it would absorb the
between-donor variance that constitutes the effect being tested. There is no
batch variable in the deposit, so nothing else is available to adjust for.

This differs from the source paper, which ran differential expression
cell-level with Seurat `FindMarkers`. That approach treats each nucleus as an
independent replicate. With thousands of nuclei per donor the effective sample
size is inflated by orders of magnitude, p-values shrink accordingly, and the
resulting gene lists reflect the number of cells sequenced as much as the
biology. Aggregating to the donor first is the standard correction.

```{r}
#| label: de

de <- cached("de_primary.rds", {
  library(DESeq2)
  out <- list()
  for (ct in sort(unique(pbmeta$cell_type))) {
    sel <- pbmeta$cell_type == ct & pbmeta$n_cells >= MIN_CELLS &
           !is.na(pbmeta$condition)
    m <- pbmeta[sel, ]
    if (any(table(m$condition) < MIN_DONORS)) next
    cts <- round(pb[, m$pseudosample, drop = FALSE]); mode(cts) <- "integer"
    cts <- cts[rowSums(cts > 0) >= ceiling(ncol(cts) / 3), , drop = FALSE]
    dds <- DESeq(DESeqDataSetFromMatrix(cts, m, ~ condition), quiet = TRUE)
    r <- as.data.frame(results(dds, contrast = c("condition", "MASH", "Control")))
    r$gene <- rownames(r); r$cell_type <- ct
    out[[ct]] <- r[order(r$padj), ]
  }
  out
})

de_severe <- cached("de_aMASH.rds", {
  library(DESeq2)
  out <- list()
  for (ct in sort(unique(pbmeta$cell_type))) {
    sel <- pbmeta$cell_type == ct & pbmeta$n_cells >= MIN_CELLS &
           pbmeta$stage %in% c("No-MASLD", "aMASH")
    m <- pbmeta[sel, ]
    m$grp <- factor(as.character(m$stage), levels = c("No-MASLD", "aMASH"))
    if (any(table(m$grp) < MIN_DONORS)) next
    cts <- round(pb[, m$pseudosample, drop = FALSE]); mode(cts) <- "integer"
    cts <- cts[rowSums(cts > 0) >= ceiling(ncol(cts) / 3), , drop = FALSE]
    dds <- DESeq(DESeqDataSetFromMatrix(cts, m, ~ grp), quiet = TRUE)
    r <- as.data.frame(results(dds, contrast = c("grp", "aMASH", "No-MASLD")))
    r$gene <- rownames(r); r$cell_type <- ct
    out[[ct]] <- r[order(r$padj), ]
  }
  out
})

deg <- data.frame(
  `Cell type` = names(de),
  Up   = sapply(de, function(r) sum(is_de(r) & r$log2FoldChange > 0)),
  Down = sapply(de, function(r) sum(is_de(r) & r$log2FoldChange < 0)),
  check.names = FALSE)
deg$Total <- deg$Up + deg$Down
deg$`vs aMASH` <- sapply(deg$`Cell type`, function(ct)
  if (ct %in% names(de_severe))
    sum(is_de(de_severe[[ct]])) else NA_integer_)
deg <- deg[order(-deg$Total), ]
kable(deg, row.names = FALSE,
      caption = paste0("Differentially expressed genes by cell type (", DE_RULE,
                       "). First three columns are MASH vs Control; the last is ",
                       "control vs advanced MASH alone (NA = too few aMASH donors to test)."))
```

Hepatocytes are by a wide margin the most transcriptionally perturbed cell type
(`r format(deg$Total[deg$"Cell type" == "Hepatocyte"], big.mark = ",")` genes at
`r DE_RULE`), consistent with the source paper's observation that hepatocytes
carry the most change with disease severity and that MASLD heritability is
enriched in them.

## Differential expression at detailed resolution

The detailed annotation subdivides hepatocytes by zonation and stellate cells by
state. The mural population identified above is relabeled here, since calling
it an inactivated stellate cell would misattribute anything it contributes.

```{r}
#| label: detailed-de

de_fine <- cached("de_detailed.rds", {
  library(BPCells); library(DESeq2)
  mt <- open_matrix_dir("bpcells_counts")
  meta <- readRDS(cache_path("meta.rds"))
  meta$ct <- as.character(meta[[COL_CT_FINE]])
  meta$ct[meta$ct == "Stellate_cell_inactivated"] <- "Mural_VSMC"
  k <- !meta$ct %in% DROP_LABELS
  grp <- paste(meta$ct[k], meta[[COL_DONOR]][k], sep = "__")
  p <- pseudobulk_matrix(mt[, which(k)], cell_groups = grp, method = "sum")
  nc <- table(grp); key <- do.call(rbind, strsplit(colnames(p), "__", fixed = TRUE))
  dm <- unique(meta[, c(COL_DONOR, COL_STAGE)]); names(dm) <- c("donor", "stage")
  m <- data.frame(pseudosample = colnames(p), cell_type = key[, 1], donor = key[, 2],
                  n_cells = as.integer(nc[colnames(p)]), stringsAsFactors = FALSE)
  m <- merge(m, dm, by = "donor"); m <- m[match(colnames(p), m$pseudosample), ]
  m$condition <- factor(unname(CONDITION_MAP[as.character(m$stage)]), levels = CONDITION_LEVELS)
  out <- list()
  for (ct in sort(unique(m$cell_type))) {
    sel <- m$cell_type == ct & m$n_cells >= MIN_CELLS & !is.na(m$condition)
    mm <- m[sel, ]; if (any(table(mm$condition) < MIN_DONORS)) next
    cts <- round(p[, mm$pseudosample, drop = FALSE]); mode(cts) <- "integer"
    cts <- cts[rowSums(cts > 0) >= ceiling(ncol(cts) / 3), , drop = FALSE]
    rownames(mm) <- mm$pseudosample
    dds <- DESeq(DESeqDataSetFromMatrix(cts, mm, ~ condition), quiet = TRUE)
    r <- as.data.frame(results(dds, contrast = c("condition", "MASH", "Control")))
    r$gene <- rownames(r); out[[ct]] <- r[order(r$padj), ]
  }
  out
})

fine_tab <- data.frame(
  `Cell type` = names(de_fine),
  DEGs = sapply(de_fine, function(r) sum(is_de(r))),
  check.names = FALSE)
kable(fine_tab[order(-fine_tab$DEGs), ], row.names = FALSE,
      caption = paste0("Differentially expressed genes (", DE_RULE,
                       "), MASH vs Control, detailed annotation."))
```

`r length(de_fine)` of the 17 detailed labels clear the donor threshold, against
10 of 12 at broad resolution. Monocytes and dendritic cells fail as before. The
mural population fails too, for a reason worth noting: it contracts so sharply
in disease that too few MASH donors retain enough of it to test. The correction
made above therefore costs nothing.

## The two axes are not the same measurement

```{r}
#| label: two-axis
#| fig-width: 7.2
#| fig-height: 5

ax <- data.frame(
  cell_type = deg$`Cell type`,
  degs      = deg$Total,
  fc        = prop_tab$`Fold change`[match(deg$`Cell type`, prop_tab$`Cell type`)],
  fdr       = prop_tab$FDR[match(deg$`Cell type`, prop_tab$`Cell type`)])
ax$abund_sig <- ifelse(ax$fdr < 0.05, "abundance FDR < 0.05", "abundance n.s.")

ggplot(ax, aes(log2(fc), degs)) +
  geom_vline(xintercept = 0, linewidth = 0.3, colour = "grey60") +
  geom_point(aes(colour = abund_sig), size = 3) +
  ggrepel::geom_text_repel(aes(label = cell_type), size = 3.1, seed = 1,
                           max.overlaps = 20, min.segment.length = 0) +
  scale_colour_manual(values = c("abundance FDR < 0.05" = "#C44E52",
                                 "abundance n.s." = "grey55"), name = NULL) +
  labs(x = "abundance change, log2(MASH / Control)",
       y = paste0("genes (", DE_RULE, ")")) +
  theme_bw(base_size = 11) + theme(legend.position = "top")
```

The two axes separate cleanly, and the separation is the substantive result.

Endothelial cells are the clearest case. Their abundance does not move at all
(fold change `r round(ax$fc[ax$cell_type == "Endothelial_cell"], 2)`, FDR
`r fmt_p(ax$fdr[ax$cell_type == "Endothelial_cell"])`), yet they carry
`r format(ax$degs[ax$cell_type == "Endothelial_cell"], big.mark = ",")`
differentially expressed genes, the second-largest program in the dataset. A
composition-only analysis would have called them unperturbed.

The reverse also holds. CD4 and CD8 T cells show the largest abundance changes
in the dataset but comparatively modest expression programs
(`r ax$degs[ax$cell_type == "CD4_T"]` and `r ax$degs[ax$cell_type == "CD8_T"]`
genes). They arrive in greater numbers without substantially altering the
program they run.

Hepatocytes are the only cell type strongly perturbed on both axes: they are
lost, and those that remain are extensively rewired.

# Synthesis across axes: which cell types are most perturbed, and how

```{r}
#| label: synthesis

top_genes <- function(ct, up = TRUE, n = 4) {
  if (!ct %in% names(de)) return("not tested")
  r <- de[[ct]]
  r <- r[is_de(r), ]
  r <- r[if (up) order(-r$log2FoldChange) else order(r$log2FoldChange), ]
  if (!nrow(r)) return("-")
  paste(head(r$gene, n), collapse = ", ")
}

syn <- data.frame(
  `Cell type` = prop_tab$`Cell type`,
  `Abundance FC` = prop_tab$`Fold change`,
  `Abundance FDR` = prop_tab$FDR,
  check.names = FALSE)
syn$DEGs <- deg$Total[match(syn$`Cell type`, deg$`Cell type`)]
syn$`Top up`   <- vapply(as.character(syn$`Cell type`), top_genes, "", up = TRUE)
syn$`Top down` <- vapply(as.character(syn$`Cell type`), top_genes, "", up = FALSE)

# Rank on both axes jointly: significant abundance change and size of the
# expression program both count toward "most perturbed".
syn$abund_rank <- rank(-abs(log2(syn$`Abundance FC`)) * (syn$`Abundance FDR` < 0.05))
syn$de_rank    <- rank(-ifelse(is.na(syn$DEGs), 0, syn$DEGs))
syn <- syn[order((syn$abund_rank + syn$de_rank) / 2), ]
syn$`Abundance FDR` <- signif(syn$`Abundance FDR`, 2)
syn$DEGs <- ifelse(is.na(syn$DEGs), "not tested", format(syn$DEGs, big.mark = ","))

kable(syn[, c("Cell type", "Abundance FC", "Abundance FDR", "DEGs",
              "Top up", "Top down")], row.names = FALSE,
      caption = paste0("Cell types ranked by combined perturbation across both axes. ",
                       "Abundance FC is MASH relative to Control; DEGs are genes at ", DE_RULE, "."))
```

Reading the table against the question asked:

**Hepatocytes are the most perturbed cell type in MASH, on both axes at once.**
They fall from `r round(stage_mean["Hepatocyte", "No-MASLD"], 1)`% to
`r round(stage_mean["Hepatocyte", "aMASH"], 1)`% of nuclei, and the survivors
carry `r format(deg$Total[deg$"Cell type" == "Hepatocyte"], big.mark = ",")`
differentially expressed genes, more than twice any other cell type. No other
population is strongly perturbed in both senses.

**Endothelial cells are the most perturbed cell type that a composition
analysis would miss entirely.** Their abundance is unchanged, and they carry the
second-largest expression program in the dataset. This is the clearest
argument for separating the two axes rather than reporting "perturbation" as one
quantity.

**T cells are the largest abundance change and a comparatively small expression
change.** CD4 and CD8 T cells expand
`r round(prop_tab$"Fold change"[prop_tab$"Cell type" == "CD4_T"], 1)`-fold and
`r round(prop_tab$"Fold change"[prop_tab$"Cell type" == "CD8_T"], 1)`-fold, with
modest programs. This is the observed evidence that they were recruited, not reprogrammed.

**Macrophages change identity rather than number.** A
`r round(prop_tab$"Fold change"[prop_tab$"Cell type" == "Macrophage"], 1)`-fold
abundance increase accompanies a wholesale switch from the homeostatic Kupffer
program to the scar-associated program, which is the most mechanistically
specific finding in the dataset.

**Stellate cells are the one case where the headline contrast is misleading.**
They are flat in abundance between Control and MASH and carry
`r format(deg$Total[deg$"Cell type" == "Stellate_cell"], big.mark = ",")`
differentially expressed genes. Both numbers understate the situation: their
expansion is confined to advanced fibrosis, and their expression program is
partly a mixture artifact of a label that contains vascular mural cells.

# What the expression changes mean

The DE results were checked against programs MASH is known to engage. These
are recovered from the data and where the data disagree
with expectation that is reported rather than smoothed over.

```{r}
#| label: programs

prog_target <- c(HSC_activation = "Stellate_cell", Scar_macrophage = "Macrophage",
                 Kupffer_homeostatic = "Macrophage", LSEC_identity = "Endothelial_cell",
                 Ductular_reaction = "Cholangiocyte")

prog <- do.call(rbind, lapply(names(prog_target), function(pn) {
  r <- de[[prog_target[pn]]]
  g <- MASH_PROGRAMS[[pn]]
  s <- r[match(g, r$gene), c("gene", "log2FoldChange", "padj")]
  data.frame(program = pn, `Cell type` = unname(prog_target[pn]),
             Gene = g, log2FC = round(s$log2FoldChange, 2),
             FDR = signif(s$padj, 2), check.names = FALSE)
}))
kable(prog, row.names = FALSE,
      caption = "Expected MASH programs, log2 fold change MASH vs Control. NA = gene did not pass the detection filter in that cell type.")
```

**The macrophage switch is recovered exactly as expected.** Scar-associated
markers rise (*SPP1*, *GPNMB*, *CD9*) while the homeostatic Kupffer program
falls (*MARCO*, *VSIG4*, *CD5L*, all FDR < 0.05). This is the documented
replacement of resident Kupffer cells by lipid- and scar-associated macrophages,
and it appears here without being sought.

*TREM2*, the canonical marker for that population, is absent from the table
because it failed the detection filter: it is non-zero in only 5 of 42
macrophage pseudo-samples. *TREM2* is present in the object, so this is not a
biological absence. Two things might be contributing: single-nucleus capture is insensitive
to it, and the deposited object was built after removing genes expressed in
fewer than 0.25% of cells in each donor, so lowly detected transcripts were
thinned before the object was assembled.

**Stellate cells appear to contradict expectation, and the contradiction is
informative.** *ACTA2* and *TAGLN* both fall significantly, the opposite of the
classical activation signature. Resolving this requires separating markers that
are specific to hepatic stellate cells from those shared with vascular mural
cells.

```{r}
#| label: stellate-resolve

hsc <- c("LRAT", "DCN", "COL1A1", "COL1A2", "COL3A1")
mural <- c("MYH11", "MYL9", "TAGLN", "ACTA2", "RGS5", "NOTCH3")
r <- de$Stellate_cell
s <- r[match(c(hsc, mural), r$gene), c("gene", "log2FoldChange", "padj")]
kable(data.frame(
  Identity = c(rep("HSC-specific", length(hsc)), rep("mural / smooth muscle", length(mural))),
  Gene = c(hsc, mural), log2FC = round(s$log2FoldChange, 2),
  FDR = signif(s$padj, 2), check.names = FALSE), row.names = FALSE,
  caption = "Stellate compartment, split by marker identity (MASH vs Control).")
```

The split is systematic. HSC-restricted genes rise (*LRAT*, the retinoid-storage
enzyme that defines hepatic stellate cells, and *DCN*), while smooth-muscle
genes fall (*MYH11*, *MYL9*, *TAGLN*). *ACTA2* is ambiguous: it marks both
activated stellate cells and vascular smooth muscle, so it cannot arbitrate.
*MYH11* and *MYL9* can, and they are unambiguously mural.

The `Stellate_cell` label therefore contains a mixture of true hepatic stellate
cells and vascular mural cells, and MASH shifts the mixture toward the former.
The apparent loss of an activation signature is a composition shift inside the
label, not deactivation of stellate cells. This is a caution about pseudobulk in
general: aggregating over a heterogeneous label can invert the direction of a
program.

**LSEC identity is only partly perturbed.** *LYVE1* falls significantly, but
*CLEC4G* and *STAB2* do not, so the classical capillarization signature is not
cleanly reproduced at this resolution despite endothelial cells carrying the
second-largest expression program in the dataset. The endothelial perturbation
is real and large but it is not well summarized by these four genes.

**The ductular reaction is not detected.** *KRT19*, *KRT7*, *SOX9* and *EPCAM*
all trend downward in cholangiocytes and none reach significance. Cholangiocyte
abundance does rise at advanced fibrosis
(`r round(stage_mean["Cholangiocyte", "No-MASLD"], 1)`% to
`r round(stage_mean["Cholangiocyte", "aMASH"], 1)`%), so any ductular expansion
here is compositional rather than a within-cell-type program, and it is
measured on the fewest donors of any tested cell type.

## Benchmarking against the reference fibrosis atlas

Ramachandran et al.
(2019) resolved the fibrotic niche of human cirrhotic liver and described three
populations that have since become the reference framework: TREM2+CD9+
scar-associated macrophages, PLVAP+/ACKR1+ scar-associated endothelial cells,
and PDGFRA+ scar-associated mesenchymal cells. All three are testable here.

The design differences matter for interpreting the result. That study profiled
end-stage cirrhotic explants by single-cell RNA-seq of dissociated whole cells.
This is needle-core biopsy tissue spanning F0 to F4, with ten donors at F3-F4,
by single-nucleus RNA-seq.

```{r}
#| label: ramachandran

niche <- list(
  `Scar-associated macrophage` = list(ct = "Macrophage",
    genes = c("TREM2","CD9","GPNMB","SPP1","FABP5","ACP5","LGALS3","CTSD")),
  `Resident Kupffer identity`  = list(ct = "Macrophage",
    genes = c("MARCO","VSIG4","CD5L","TIMD4","LYVE1","CETP")),
  `Scar-associated endothelium` = list(ct = "Endothelial_cell",
    genes = c("PLVAP","ACKR1","CD34","VWA1")),
  `LSEC identity`              = list(ct = "Endothelial_cell",
    genes = c("CLEC4G","CLEC4M","STAB2","OIT3","FCN2","LYVE1")),
  `PDGFRA+ mesenchyme`         = list(ct = "Stellate_cell",
    genes = c("PDGFRA","OSMR","COL1A1","COL3A1","LOXL1")))

ram <- do.call(rbind, lapply(names(niche), function(nm) {
  x <- niche[[nm]]; r <- de[[x$ct]]; i <- match(x$genes, r$gene)
  data.frame(program = nm, `Cell type` = x$ct, Gene = x$genes,
             log2FC = round(r$log2FoldChange[i], 2),
             FDR = signif(r$padj[i], 2), check.names = FALSE)
}))
ram$Result <- ifelse(is.na(ram$FDR), "not detected",
              ifelse(ram$FDR >= FDR_CUTOFF | abs(ram$log2FC) < LFC_CUTOFF, "n.s.",
              ifelse(ram$log2FC > 0, "up", "down")))
kable(ram, row.names = FALSE,
      caption = paste0("Ramachandran et al. fibrotic-niche programs, tested in this ",
                       "dataset (MASH vs Control). Result applies ", DE_RULE, "."))
```

**The scar-associated macrophage population replicates cleanly.** *SPP1*,
*FABP5*, *ACP5*, *GPNMB*, *LGALS3* and *CTSD* all rise significantly, with *CD9*
just short of threshold and *TREM2* undetectable for the reasons given in the
limitations. The reciprocal half is equally clear: the resident Kupffer
program falls across every marker tested.

*TIMD4* is the informative one. It marks embryonically derived, self-renewing
Kupffer cells, and its loss
(`r ram$log2FC[ram$Gene == "TIMD4"]`, FDR `r fmt_p(ram$FDR[ram$Gene == "TIMD4"])`)
alongside gain of the TREM2/CD9 program is the signature of resident
macrophages being replaced by monocyte-derived cells. That is the mechanism
Ramachandran described in cirrhosis, recovered here in a different disease,
tissue source and assay.

**The scar-associated endothelial population is not reproduced, and the data
point the other way.** *PLVAP* falls rather than rises. Because pseudobulk over
all endothelial cells could dilute a rare subpopulation, the fraction of
endothelial nuclei expressing it was tested directly at donor level.

```{r}
#| label: plvap-fraction

plvap <- cached("plvap_fraction.rds", {
  library(BPCells)
  mt <- open_matrix_dir("bpcells_counts")
  meta <- readRDS(cache_path("meta.rds"))
  sel <- meta[[COL_CT_BROAD]] == "Endothelial_cell"
  v <- as.vector(as.matrix(mt["PLVAP", which(sel)]))
  data.frame(donor = meta[[COL_DONOR]][sel], stage = meta[[COL_STAGE]][sel],
             pos = v > 0)
})

pf <- aggregate(pos ~ donor, plvap, function(x) mean(x) * 100)
pf$condition <- donors$condition[match(pf$donor, donors$donor)]
tt <- t.test(pos ~ condition, data = subset(pf, !is.na(condition)))

kable(data.frame(
  Group = c("Control", "MASH"),
  `PLVAP+ endothelial nuclei (%)` = round(unname(tt$estimate), 2),
  check.names = FALSE), row.names = FALSE,
  caption = paste0("Donor-level fraction of PLVAP-expressing endothelial nuclei (p = ",
                   signif(tt$p.value, 2), ")."))
```

PLVAP-positive endothelial cells are uncommon in this tissue and become
significantly *less* common in disease, falling from
`r round(tt$estimate[1], 1)`% to `r round(tt$estimate[2], 1)`% of endothelial
nuclei (p `r fmt_p(tt$p.value)`). *ACKR1* is present in the object but too
sparsely detected to test.

There is also no capillarization. Sinusoidal identity is preserved across the
whole progression: *STAB2*, *CLEC4G*, *CLEC4M*, *OIT3* and *FCN2* are unchanged
or slightly up, and only *LYVE1* declines. That is consistent with the
endothelial co-expression module reported below, `endo-M8`, which carries
*STAB2*, *OIT3* and *CLEC4M* and moves upward. Endothelial cells in this cohort
are transcriptionally very active, as the
`r format(sum(is_de(de$Endothelial_cell)), big.mark = ",")` differentially expressed genes
show, but the direction of that activity is inflammatory rather than a loss of
sinusoidal identity.

**The PDGFRA+ mesenchymal population is not reproduced either.** *PDGFRA* and
the collagens do not move significantly. This is the same gap as the absent
myofibroblast compartment documented in the limitations, seen from another
angle. The one suggestion of the program is *OSMR*, which is not significant
in the primary contrast but is in the advanced-fibrosis contrast.

```{r}
#| label: osmr
osmr <- de_severe$Stellate_cell[de_severe$Stellate_cell$gene == "OSMR",
                                c("log2FoldChange", "padj")]
kable(data.frame(Gene = "OSMR", `Cell type` = "Stellate_cell",
                 Contrast = "control vs aMASH",
                 log2FC = round(osmr$log2FoldChange, 2),
                 FDR = signif(osmr$padj, 2), check.names = FALSE),
      row.names = FALSE)
```


### What the pattern of hits and misses says

The two failures are not independent. Both scar-associated endothelium and
PDGFRA+ mesenchyme are components of a fibrotic niche organized around
myofibroblasts and scar matrix, and both are physically embedded in it. The
population that does replicate, scar-associated macrophages, is the one that is
not. Combined with the absent myofibroblast compartment, I think the
explanation is that this tissue preparation does not recover cells resident in
dense scar, rather than that the niche is absent from MASH.

Disease stage points the same way. The reference study profiled end-stage
cirrhosis while the most advanced donors here are F3-F4 and there are ten of them.

This makes a concrete prediction, which is the useful form for a negative
result: a MASH dataset built from precision-cut slices, resections or explants
should recover the scar-associated endothelial and mesenchymal populations,
while a needle-core study of comparable stage should not. What can be said from
this cohort is that the macrophage arm of the fibrotic niche is present and
measurable in MASH well before cirrhosis.

# Pathway-level interpretation

The method I employed is GSEA (`fgsea`) on the MSigDB Hallmark collection, ranking genes by
the DESeq2 Wald statistic. GSEA rather than over-representation testing because
the pseudobulk DE produces a signed statistic for every tested gene, and ranking
on all of it uses information that a hard FDR < 0.05 cutoff would discard.
Hallmark rather than a finer collection because the question is what is
happening to each cell type, and Hallmark's 50 sets are curated to minimize the
redundancy that makes larger collections unreadable.

```{r}
#| label: gsea

gsea <- cached("gsea_primary.rds", {
  library(fgsea); library(msigdbr)
  hall <- msigdbr(species = "Homo sapiens", collection = "H")
  sets <- split(hall$gene_symbol, hall$gs_name)
  names(sets) <- sub("^HALLMARK_", "", names(sets))
  out <- list()
  for (ct in names(de)) {
    r <- de[[ct]]; r <- r[!is.na(r$stat), ]
    stats <- setNames(r$stat, r$gene); stats <- stats[!duplicated(names(stats))]
    set.seed(42)
    res <- fgsea(sets, stats, minSize = 10, maxSize = 500, nPermSimple = 10000)
    res$cell_type <- ct
    out[[ct]] <- res[order(res$padj), ]
  }
  out
})

gs_all <- do.call(rbind, lapply(names(gsea), function(ct) {
  d <- as.data.frame(gsea[[ct]]); d[, c("pathway", "NES", "padj", "cell_type")]
}))
```

## What each cell type is doing

```{r}
#| label: gsea-heatmap
#| fig-width: 9
#| fig-height: 7
#| column: page

# Restrict to pathways that are significant somewhere with a substantial effect.
# Every significant set would be 44 rows, which is unreadable at display size;
# this keeps the strongest 25 and the panel legible.
keep_p <- unique(subset(gs_all, padj < 0.05 & abs(NES) > 1.75)$pathway)
hm <- subset(gs_all, pathway %in% keep_p)
hm$label <- ifelse(hm$padj < 0.05, "*", "")
hm$NES_show <- ifelse(hm$padj < 0.05, hm$NES, NA)

ord <- tapply(hm$NES_show, hm$pathway, function(x) mean(x, na.rm = TRUE))
hm$pathway <- factor(hm$pathway, levels = names(sort(ord)))

ggplot(hm, aes(cell_type, pathway, fill = NES_show)) +
  geom_tile(colour = "grey92", linewidth = 0.3) +
  geom_text(aes(label = label), size = 3, vjust = 0.75) +
  scale_fill_gradient2(low = "#3B4CC0", mid = "white", high = "#B40426",
                       midpoint = 0, na.value = "grey96", name = "NES") +
  labs(x = NULL, y = NULL,
       caption = "* FDR < 0.05. Grey = not significant in that cell type.") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 40, hjust = 1),
        axis.text.y = element_text(size = 9),
        panel.grid = element_blank())
```

**Hepatocytes lose their specialized metabolic identity and mount an injury
response.** Bile acid metabolism, xenobiotic metabolism and fatty acid
metabolism all fall, while the p53 pathway rises
(NES `r round(subset(gs_all, cell_type == "Hepatocyte" & pathway == "P53_PATHWAY")$NES, 2)`,
FDR `r fmt_p(subset(gs_all, cell_type == "Hepatocyte" & pathway == "P53_PATHWAY")$padj)`)
alongside all three proliferation sets. That combination, loss of differentiated
function with simultaneous stress signalling and cell-cycle entry, is the
transcriptional signature of a compartment being damaged and regenerating. It
independently reproduces the source paper's description of reduced hepatocyte
metabolic function in MASH.

**Macrophages turn on cholesterol handling.** `CHOLESTEROL_HOMEOSTASIS` is the
strongest positive result in that cell type
(NES `r round(subset(gs_all, cell_type == "Macrophage" & pathway == "CHOLESTEROL_HOMEOSTASIS")$NES, 2)`),
with TNFA/NF-kB signalling also up. This is the pathway-level counterpart of the
gene-level switch reported above: the *GPNMB*, *SPP1* and *CD9* population that
replaces homeostatic Kupffer cells is a lipid-handling, inflammatory macrophage.

**Endothelial cells are inflammatory, not merely present.** `TNFA_SIGNALING_VIA_NFKB`
is their top result
(NES `r round(subset(gs_all, cell_type == "Endothelial_cell" & pathway == "TNFA_SIGNALING_VIA_NFKB")$NES, 2)`).
This supplies the missing interpretation for the most striking result in the
abundance analysis: endothelial cells do not change in number at all, yet carry
the second-largest expression program in the dataset, and that program is
inflammatory activation.

**Cholangiocytes activate Hedgehog signalling**
(NES `r round(subset(gs_all, cell_type == "Cholangiocyte" & pathway == "HEDGEHOG_SIGNALING")$NES, 2)`)
together with inflammatory response, while oxidative phosphorylation falls.
Hedgehog is an established fibrogenic pathway in MASH and is consistent with the
cholangiocyte expansion seen at advanced fibrosis, even though the classical
*KRT19*/*KRT7* ductular markers did not reach significance gene by gene.

**Proliferation moves in opposite directions, which is the clearest evidence the
results are biological.** Hepatocytes gain E2F targets, G2M checkpoint and
mitotic spindle while NK cells lose all three, and lose them strongly
(`E2F_TARGETS` NES `r round(subset(gs_all, cell_type == "NK" & pathway == "E2F_TARGETS")$NES, 2)`,
FDR `r fmt_p(subset(gs_all, cell_type == "NK" & pathway == "E2F_TARGETS")$padj)`).
The lymphoid populations that expand several-fold in abundance are therefore not
expanding by local proliferation. That is consistent with recruitment, and with
the earlier observation that T cells change greatly in number while changing
their expression program comparatively little. It is consistent with
recruitment rather than a demonstration of it. That said, this dataset is a single snapshot
per donor and cannot establish where those cells came from.

# Resolving the signature

## Is the hepatocyte response zonal?

Hepatocytes split into pericentral, periportal and interzonal, and the analysis so far appears to say the pericentral zone is most perturbed. However, that reading does
not survive scrutiny as the zones differ in donor coverage and in cell numbers, so
they differ in power before any biology is considered.

```{r}
#| label: zonation

z <- c("Hepatocyte_pericentral", "Hepatocyte_periportal", "Hepatocyte_interzonal")
lab <- c("PC", "PP", "IZ")
cz <- Reduce(intersect, lapply(de_fine[z], function(r) r$gene))
fz <- sapply(de_fine[z], function(r) r$log2FoldChange[match(cz, r$gene)])
az <- sapply(de_fine[z], function(r) r$padj[match(cz, r$gene)])
colnames(fz) <- colnames(az) <- lab
sigz <- !is.na(az) & az < FDR_CUTOFF & abs(fz) >= LFC_CUTOFF
shared <- rowSums(sigz) == 3
concordant <- sum(abs(rowSums(sign(fz[shared, ]))) == 3, na.rm = TRUE)

# Donors contributing a testable pseudo-sample, counted the same way the DE did.
zone_donors <- sapply(z, function(x) {
  k <- md[[COL_CT_FINE]] == x
  n <- table(md[[COL_DONOR]][k])
  d <- unique(md[, c(COL_DONOR, "condition")])
  d <- d[match(names(n), d[[COL_DONOR]]), ]
  ok <- n >= MIN_CELLS & !is.na(d$condition)
  paste(sum(ok & d$condition == "Control"), "/", sum(ok & d$condition == "MASH"))
})

kable(data.frame(
  Zone = lab,
  `Donors (Control / MASH)` = zone_donors,
  `Nuclei` = format(sapply(z, function(x) sum(md[[COL_CT_FINE]] == x)), big.mark = ","),
  `DEGs` = sapply(z, function(x) sum(is_de(de_fine[[x]]))),
  `Median |log2FC|` = round(apply(abs(fz), 2, median, na.rm = TRUE), 3),
  check.names = FALSE), row.names = FALSE,
  caption = "Hepatocyte zones: differential expression against available power.")
```

The DEG counts track donor and nucleus numbers, not effect size. Median absolute
fold change is essentially identical across the three zones
(`r paste(round(apply(abs(fz), 2, median, na.rm = TRUE), 2), collapse = ", ")`),
and the pairwise correlation of fold changes between zones is
`r round(min(cor(fz, use = "complete.obs")[lower.tri(diag(3))]), 2)` to
`r round(max(cor(fz, use = "complete.obs")[lower.tri(diag(3))]), 2)`.

Of the `r format(sum(shared), big.mark = ",")` genes significant in all three
zones, `r format(concordant, big.mark = ",")` agree in direction. That is
`r round(100 * concordant / sum(shared))`%.

**Hepatocytes mount one disease program, not three zonal ones.** This matters
for how the result is used. If the response were zone-restricted, any
intervention would have to reach the right zone to be effective. It is not:
zonal identity is preserved as a baseline, and the MASH response is superimposed
on all three compartments in the same direction and at the same magnitude.

## Module-level structure

Individual genes are a difficult unit to act on. The source study built
co-expression modules with hdWGCNA and deposited both the module definitions and
per-cell harmonized module eigengenes. Those are reused here rather than
recomputed, but tested at the donor level, which their own analysis did not do.

```{r}
#| label: modules

mod_res <- cached("module_de.rds", {
  dirp <- "data/unzipped/Zenodo/hdWGCNA_results"
  meta <- readRDS(cache_path("meta.rds"))
  meta$cond <- unname(CONDITION_MAP[meta[[COL_STAGE]]])
  do.call(rbind, lapply(c("hepatocyte","stellate_cell","endothelial_cell","cholangiocyte"),
    function(tag) {
      x <- read.delim(gzfile(file.path(dirp, paste0(tag, "_hME.txt.gz"))), check.names = FALSE)
      rownames(x) <- x[[1]]; x <- x[, -1, drop = FALSE]
      x <- x[, colnames(x) != "grey", drop = FALSE]
      cm <- intersect(rownames(x), rownames(meta))
      x <- x[cm, , drop = FALSE]; mm <- meta[cm, ]
      dmean <- t(sapply(colnames(x), function(k)
        tapply(x[[k]], mm[[COL_DONOR]], mean, na.rm = TRUE)))
      dn <- unique(mm[, c(COL_DONOR, "cond")]); rownames(dn) <- dn[[COL_DONOR]]
      dn <- dn[colnames(dmean), ]; kp <- !is.na(dn$cond)
      r <- do.call(rbind, lapply(rownames(dmean), function(k) {
        tt <- t.test(dmean[k, kp] ~ factor(dn$cond[kp], levels = CONDITION_LEVELS))
        data.frame(cell_type = tag, module = k,
                   diff = unname(tt$estimate[2] - tt$estimate[1]), p = tt$p.value)
      }))
      r$FDR <- p.adjust(r$p, "BH"); r
    }))
})

sig_mod <- subset(mod_res, FDR < 0.05)
sig_mod <- sig_mod[order(sig_mod$FDR), ]
kable(data.frame(
  `Cell type` = sig_mod$cell_type, Module = sig_mod$module,
  `Change in MASH` = round(sig_mod$diff, 2), FDR = signif(sig_mod$FDR, 2),
  check.names = FALSE), row.names = FALSE,
  caption = "Co-expression modules differentially expressed between MASH and Control, tested on per-donor mean module eigengenes.")
```

`r nrow(sig_mod)` of `r nrow(mod_res)` deposited modules move significantly at
donor level. The distribution matches the gene-level results closely:
`r sum(sig_mod$cell_type == "endothelial_cell")` of nine endothelial modules
shift despite endothelial cells not changing in abundance at all, which is the
same dissociation reported earlier arriving through an independent construction.

The largest hepatocyte effect is **Hep-M12**, down by
`r round(abs(sig_mod$diff[sig_mod$module == "Hep-M12"]), 2)`
(FDR `r fmt_p(sig_mod$FDR[sig_mod$module == "Hep-M12"])`). Its highest-connectivity
members are *FGB*, *FGG*, *FGA*, *FGL1*, *CP*, *CYP3A5* and *PCK1*: fibrinogen
synthesis, copper transport, drug metabolism and gluconeogenesis. It is a
hepatocyte secretory and metabolic identity program, and it is the single
most-perturbed coordinated unit found anywhere in this analysis.

Two further observations: The upregulated
hepatocyte module Hep-M6 contains *PPARA*, *MLXIPL* and *PLIN5*, and the
upregulated stellate module HSC-M3 contains *NR1H4*. Both are nuclear receptors
with existing pharmacology in metabolic liver disease, which is a reassuring
sign that the signature is not disconnected from known biology. Nothing here directly
tests any drug, and no directional inference about a therapeutic effect follows
from a module changing.

# Discussion

## What this dataset can and cannot say about MASH therapeutics

MASH now has FDA approved drugs and a crowded development pipeline, and a natural
question is whether a dataset like this speaks to any of them. It does, but only
in a specific and limited way.

An observational cross-sectional cohort of 48 donors can establish two things
about a drug target: whether the target is expressed in the cell type the drug
is meant to act on, and whether the program that target controls is among the
things disease has perturbed. Those are questions about whether a mechanism is
*engageable* and *relevant* in human MASH liver. 

With that stated, the dataset does separate the pipeline into three groups.

```{r}
#| label: drug-targets

drug_expr <- cached("drug_target_expr.rds", {
  library(BPCells)
  mt <- open_matrix_dir("bpcells_counts")
  meta <- readRDS(cache_path("meta.rds"))
  k <- !meta[[COL_CT_BROAD]] %in% DROP_LABELS
  ctf <- factor(as.character(meta[[COL_CT_BROAD]])[k]); sb <- mt[, which(k)]
  nmz <- log1p(multiply_cols(sb, 1e4 / colSums(sb)))
  gg <- c("THRB","DIO1","GLP1R","GCGR","GIPR","KLB","FGFR1","NR1H4","NR0B2",
          "PPARA","PPARD","PPARG","ITGAV","ITGB1","ITGB6","ITGB8",
          "TGFB1","TGFBR1","TGFBR2","LOXL2","LOX","LOXL1")
  present <- gg[gg %in% rownames(mt)]
  a <- pseudobulk_matrix(nmz[present, ], cell_groups = ctf, method = "mean")
  n <- pseudobulk_matrix(sb[present, ],  cell_groups = ctf, method = "nonzeros")
  list(avg = a, pct = sweep(n, 2, as.integer(table(ctf)), "/") * 100,
       absent = setdiff(gg, present))
})

targ <- data.frame(
  Target = c("THRB", "GCGR", "GLP1R", "GIPR", "KLB", "NR1H4", "PPARA", "PPARD",
             "PPARG", "ITGAV", "ITGB6", "TGFB1", "LOXL2", "LOX"),
  Drug = c("resmetirom", "survodutide (glucagon arm)", "semaglutide",
           "tirzepatide (GIP arm)", "FGF21 analogues", "FXR agonists",
           "elafibranor, lanifibranor", "seladelpar, elafibranor",
           "lanifibranor", "integrin inhibitors", "integrin avb6 inhibitors",
           "TGF-beta blockade", "simtuzumab", "LOX inhibitors"),
  stringsAsFactors = FALSE)

look <- function(g, what) {
  if (g %in% drug_expr$absent) return(NA_real_)
  m <- drug_expr[[what]]
  if (!g %in% rownames(m)) return(NA_real_)
  max(m[g, ])
}
where <- function(g) {
  if (g %in% drug_expr$absent) return("not in object")
  colnames(drug_expr$avg)[which.max(drug_expr$avg[g, ])]
}
targ$`Highest in`  <- sapply(targ$Target, where)
targ$`Mean expr`   <- round(sapply(targ$Target, look, "avg"), 2)
targ$`% expressing`<- round(sapply(targ$Target, look, "pct"), 1)
kable(targ, row.names = FALSE,
      caption = "Where each target is expressed. Mean expression and percent are for the cell type where the target is highest.")
```

### Targets that are well represented in this tissue

**THR-beta (resmetirom).** *THRB* is expressed in
`r round(drug_expr$pct["THRB", "Hepatocyte"], 1)`% of hepatocytes, the highest of
any cell type, and it is not itself altered in disease. The programs it
regulates are hepatocyte fatty acid metabolism, bile acid metabolism and
xenobiotic metabolism are all significantly depleted in MASH here, and that loss
of differentiated metabolic function is the dominant hepatocyte finding in this
report. So the receptor is abundant and intact in exactly the cell type whose
metabolic program has degraded. That is a coherent target-tissue picture, and
it is the most straightforwardly supported mechanism in the dataset.

**FXR (obeticholic acid and successors).** *NR1H4* is well expressed in both
hepatocytes and stellate cells. More informative than the receptor is its
output: *NR0B2*, the canonical FXR target gene, is strongly reduced in
hepatocytes
(log2FC `r round(de$Hepatocyte$log2FoldChange[de$Hepatocyte$gene == "NR0B2"], 2)`,
FDR `r fmt_p(de$Hepatocyte$padj[de$Hepatocyte$gene == "NR0B2"])`), alongside the
depleted bile acid program. FXR signalling output is measurably lower in MASH
hepatocytes in this cohort. Whether restoring it is beneficial is a clinical
question this dataset cannot address, and the mixed clinical history of FXR
agonism is a reminder that pathway restoration and patient benefit are different
things.

**PPAR-alpha (elafibranor, lanifibranor).** *PPARA* is well expressed in
hepatocytes and unchanged in disease, and it sits inside `Hep-M6`, one of the
hepatocyte modules that moves significantly. *PPARG* is macrophage-biased, and
macrophages show the strongest cholesterol-homeostasis enrichment in the
dataset, which is at least consistent with a lipid-handling role for the
myeloid arm of a pan-PPAR agent. *PPARD* is poorly detected in every cell type
here, so this dataset can say little about selective delta agonism.

**Glucagon (survodutide's glucagon arm).** This one differs, and the distinction
is worth drawing. *GCGR* is expressed in
`r round(drug_expr$pct["GCGR", "Hepatocyte"], 1)`% of hepatocytes and is
modestly reduced in MASH
(log2FC `r round(de$Hepatocyte$log2FoldChange[de$Hepatocyte$gene == "GCGR"], 2)`,
FDR `r fmt_p(de$Hepatocyte$padj[de$Hepatocyte$gene == "GCGR"])`). Unlike the
GLP-1 and GIP components, the glucagon arm of a dual agonist has a direct
hepatocyte target that is measurable in this tissue. That makes glucagon-containing
dual agonists the incretin-class agents this dataset can actually say something
about.

### Targets whose liver action is not visible in this dataset

**GLP-1 (semaglutide).** *GLP1R* is not present in the object at all. This is not
a detection failure peculiar to this study: hepatocytes are not generally held to
express GLP-1 receptor, and the hepatic benefit of GLP-1 agonism is understood to
be largely indirect, through weight loss, glycaemic control and adipose
signalling. The consequence is simply that a liver transcriptomic atlas is the
wrong instrument for evaluating it. Its absence here is uninformative about the
drug.

**GIP (tirzepatide's GIP arm).** *GIPR* is effectively undetected in hepatocytes.
The same reasoning applies.

**FGF21 analogues.** The obligate co-receptor *KLB* is detected in only
`r round(drug_expr$pct["KLB", "Hepatocyte"], 1)`% of hepatocytes, and *FGFR1* is
essentially absent from hepatocytes while being present on stellate cells. FGF21
biology is substantially adipose-directed, so a liver dataset again under-reports
the mechanism.

### Anti-fibrotic targets: this dataset cannot evaluate them

```{r}
#| label: antifibrotic-targets

af <- c("LOX", "LOXL1", "LOXL2", "ITGB6", "ITGB1", "ITGAV", "TGFB1", "TGFBR1", "TGFBR2")
af <- af[!af %in% drug_expr$absent]
kable(data.frame(
  Target = af,
  `Highest in` = sapply(af, where),
  `Mean expr` = round(sapply(af, look, "avg"), 2),
  `% expressing` = round(sapply(af, look, "pct"), 1),
  check.names = FALSE), row.names = FALSE,
  caption = "Anti-fibrotic drug targets, at the cell type where each is highest.")
```

The lysyl oxidases are the clearest case. *LOX* is essentially undetected,
*LOXL1* and *LOXL2* are detected in a few percent of stellate cells, and *LOXL2*
does not change significantly with disease. *ITGB6*, the target of alpha-v-beta-6
integrin inhibition, is effectively absent. *TGFB1* is low in every cell type,
and the TGF-beta signalling program is not significantly enriched in stellate
cells. Read naively, this would look like a dataset arguing against anti-fibrotic
targets.

That reading would be wrong, and the reason is documented earlier in this report.
These drugs act on collagen-producing myofibroblasts, and this dataset does not
contain them: the highest *COL1A1* count in any nucleus across the entire atlas
is
`r max(col1a1_counts())`, and the scar-associated
endothelial and mesenchymal populations from the reference cirrhosis atlas both
fail to appear while the scar-associated macrophage population replicates
cleanly. The most likely explanation for all of it is that needle-core
biopsy tissue does not recover cells resident in dense scar matrix.

So the correct statement is not that this dataset provides weak support for
anti-fibrotic mechanisms. It is that **this dataset cannot assess them**, because
the cell population those drugs target is missing from it for reasons of tissue
sampling rather than biology. An apparent absence of a drug target is only
evidence when the cells that would express it were recoverable in the first
place. The one anti-fibrotic-adjacent target that does move is *ITGAV* in
stellate cells
(log2FC `r round(de$Stellate_cell$log2FoldChange[de$Stellate_cell$gene == "ITGAV"], 2)`,
FDR `r fmt_p(de$Stellate_cell$padj[de$Stellate_cell$gene == "ITGAV"])`), which is
suggestive and no more.

### Does the data support combination therapy?

Not directly, and any dataset of this design would struggle to. What it does
provide is a structural argument, which is weaker but not empty.

The central result of this report is that MASH perturbation is distributed
across cell types along two independent axes, and that the axes do not coincide.
Hepatocytes lose a metabolic program and are depleted. Macrophages switch
identity from resident Kupffer to scar-associated without much change in number.
Endothelial cells carry the second-largest expression program in the dataset
with no abundance change at all, and its content is inflammatory. Lymphocytes
expand several-fold with comparatively little transcriptional change.

A hepatocyte-directed metabolic agent, whether a thyromimetic, an incretin or a
PPAR agonist, addresses the first of those compartments. Nothing in this data
suggests it would directly reach the macrophage identity switch or the
endothelial inflammatory program, both of which are large, statistically
robust and confined to cell types the metabolic targets are not expressed in.
That distribution is a reason to expect that a single-compartment intervention
leaves a substantial part of the measured disease untouched.

What this dataset cannot do is tell you which second agent to employ. Adjudicating a
metabolic-plus-anti-fibrotic combination specifically would require measuring
the fibrotic compartment, and this tissue preparation does not recover it (but at least we now know that). A
study built on precision-cut liver slices (like the kind I've done in the past at Pliant), resections or explants, where
myofibroblasts and the scar-resident niche are retained, is the design that
could potentially answer the direct anti-fibrotic question.

## Limitations

**Compositional dependence.** Proportions sum to one, so the loss of half the
hepatocyte compartment mechanically inflates every other fraction. The immune
expansions are relative enrichments. Distinguishing that from absolute
infiltration would require a cell-number-per-unit-tissue measurement that
dissociation-based data cannot provide.

**No batch variable.** The deposit carries no sequencing run, chemistry or
processing-date column, so no technical covariate could be tested or adjusted
for. The cohort is single-center, which limits the risk, but this cannot be
verified from what was deposited.

**Fibrosis is available only as a dichotomy.** The source study staged fibrosis
on the full Brunt/Kleiner F0-F4 scale, but used it only to define the eMASH
(F0-2) and aMASH (F3-4) groups, and the per-donor stage was not deposited. The
metadata carries disease group and nothing else clinical: no fibrosis stage, no
activity score, no ALT, and no age, sex or BMI. The finest fibrosis comparison
possible here is therefore the two-bin eMASH-versus-aMASH contrast, on 9 and 10
donors. No dose-response across F0 to F4 can be tested, and the stellate result
above should be read with that in mind: it is a difference between two coarse
fibrosis bins, not a gradient. The same constraint rules out relating cell-type
proportions to clinical severity, which the source paper was able to do because
it had the clinical table.

**Label heterogeneity is a real constraint.** The stellate result demonstrates
that pseudobulk over a mixed label can invert a program's apparent direction.
The same risk applies in principle to every label used here. It was caught in
this instance because the expected direction was known in advance and checked.

**Controls are not healthy volunteers.** They are biopsies from people
undergoing clinical procedures with no MASLD diagnosis. A baseline of stellate
activation and immune presence is expected, and comparisons are against that
baseline, not against normal liver.

**Unequal and modest group sizes.** The stage-resolved contrasts rest on 6 to 10
donors per arm. The eMASH-versus-aMASH stellate result does not survive
multiple-testing correction and is reported as suggestive for that reason.

**Reused integration and inherited annotation.** The analysis relies on the
deposited Liger integration and cluster labels rather than re-deriving them. That is deliberate: the labels
were validated against canonical markers before use, and re-clustering a
published, already-integrated object risks recapitulating the original result
rather than testing it. It does mean any error in the deposited annotation
propagates here, as the stellate `inactivated` subcluster illustrates. That case
also shows the check working: the mixture was detectable from marker expression,
from unbiased enrichment and from the subcluster-level profile, without
re-clustering anything.


# Supplementary

## Within-lineage resolution

Two lineages subdivide in the deposited annotation, and both subdivide along an
axis MASH acts on. This is where the detailed labels earn their place.

```{r}
#| label: detailed
#| fig-width: 10
#| fig-height: 3.6

sub_frac <- function(prefix) {
  s  <- md[grepl(prefix, md[[COL_CT_FINE]]), ]
  tb <- table(s[[COL_DONOR]], s[[COL_CT_FINE]])
  pr <- sweep(tb, 1, rowSums(tb), "/") * 100
  st <- donors$stage[match(rownames(pr), donors$donor)]
  data.frame(donor = rep(rownames(pr), ncol(pr)),
             subtype = rep(colnames(pr), each = nrow(pr)),
             pct = as.vector(pr), stage = rep(st, ncol(pr)))
}

st_frac  <- sub_frac("^Stellate")
hep_frac <- sub_frac("^Hepatocyte")

ggplot(rbind(st_frac, hep_frac), aes(stage, pct, fill = stage)) +
  geom_boxplot(outlier.shape = NA, linewidth = 0.3) +
  geom_jitter(width = 0.15, size = 0.6, alpha = 0.5) +
  facet_wrap(~ subtype, nrow = 2, scales = "free_y") +
  scale_fill_manual(values = c("#4C72B0", "#DD8452", "#C44E52", "#8172B3")) +
  labs(x = NULL, y = "% of parent lineage") +
  theme_bw(base_size = 9) +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 30, hjust = 1))
```

```{r}
#| label: detailed-table
sm <- function(d) round(sapply(levels(donors$stage), function(s)
  tapply(d$pct[d$stage == s], d$subtype[d$stage == s], mean, na.rm = TRUE)), 1)
kable(rbind(sm(st_frac), sm(hep_frac)),
      caption = "Mean composition within each lineage (%), by stage.")
```

Hepatocyte zonation shifts consistently: the pericentral fraction rises across
the progression while periportal and interzonal fractions fall. Given that
hepatocytes are simultaneously halving in absolute proportion, this is a
relative change in which zonal identity survives, not evidence that pericentral
hepatocytes increase in number.

The stellate substates move much less than their names suggest. The activated
fraction is already
`r round(mean(st_frac$pct[st_frac$subtype == "Stellate_cell_activated" & st_frac$stage == "No-MASLD"], na.rm = TRUE), 1)`%
in controls and reaches only
`r round(mean(st_frac$pct[st_frac$subtype == "Stellate_cell_activated" & st_frac$stage == "aMASH"], na.rm = TRUE), 1)`%
in advanced MASH. The controls are liver biopsies from people undergoing
clinical procedures rather than healthy volunteers, so a non-trivial baseline of
stellate activation is plausible.

The substate that does move sharply is `inactivated`, which falls from
`r round(mean(st_frac$pct[st_frac$subtype == "Stellate_cell_inactivated" & st_frac$stage == "No-MASLD"], na.rm = TRUE), 1)`%
to
`r round(mean(st_frac$pct[st_frac$subtype == "Stellate_cell_inactivated" & st_frac$stage == "aMASH"], na.rm = TRUE), 1)`%.
That turns out to matter more than the activated fraction, and it is taken up in
detail below, after the enrichment results supply a second line of evidence
about what that subcluster actually contains.

How these substate labels were produced is relevant to reading them. The source
paper's supplementary methods describe subclustering all donors together, then
subclustering disease and control cells separately, and assigning a substate
only where both analyses agreed. Cells that disagreed were labeled
`Unidentified`, and subclusters showing ectopic marker expression were labeled
`misc`. That is a conservative procedure, but it means the substate assignments
are not independent of disease status, so substate proportions compared across
disease groups should be read as descriptive rather than as a clean test.


## Is this a technical gradient?

Enrichment on pseudobulk is vulnerable to a specific failure: proliferation and
biosynthesis sets are loaded with highly expressed housekeeping genes, so a
systematic difference in capture depth between groups can produce a sweep of
significant hits that reflect nothing biological. Three checks decide whether
that is happening here.

```{r}
#| label: gsea-diagnostic

sig <- subset(gs_all, padj < 0.05)
spread <- aggregate(NES ~ pathway, sig, function(x) c(n = length(x), lo = min(x), hi = max(x)))
spread <- data.frame(pathway = spread$pathway, n_types = spread$NES[, "n"],
                     NES_min = round(spread$NES[, "lo"], 2),
                     NES_max = round(spread$NES[, "hi"], 2))
spread <- spread[order(-spread$n_types), ][1:6, ]

pb_sel <- pbmeta$n_cells >= MIN_CELLS & !is.na(pbmeta$condition)
depth  <- tapply(colSums(pb[, pbmeta$pseudosample[pb_sel]]) /
                 pbmeta$n_cells[pb_sel], pbmeta$condition[pb_sel], median)

kable(spread, row.names = FALSE,
      caption = "Pathways significant in the most cell types, with the range of their enrichment scores.")
```

**The most widely significant pathways change sign between cell types.**
`MYC_TARGETS_V1` runs from NES `r round(min(subset(sig, pathway == "MYC_TARGETS_V1")$NES), 2)`
to `r round(max(subset(sig, pathway == "MYC_TARGETS_V1")$NES), 2)` depending on
which cell type is examined. A depth artifact would push every cell type in the
same direction.

**Median counts per nucleus are comparable between groups**
(`r round(depth[["Control"]])` in Control, `r round(depth[["MASH"]])` in MASH),
and DESeq2 models library size explicitly in any case.

**The leading-edge genes are process-specific, not housekeeping.** The strongest
single result in the dataset is loss of `E2F_TARGETS` in NK cells, and its
leading edge is *RRM2*, *MELK*, *BIRC5*, *MCM4*, *MCM6*, *ORC6*, *CENPE*:
replication and mitotic machinery, with no ribosomal protein genes at all.

The results are therefore read as biological.

## Localizing the mixture to one subcluster

The annotation section established that the subcluster labeled
`Stellate_cell_inactivated` is vascular mural rather than stellate. This section
works out what that mislabel does to the aggregate stellate results, which is
where it matters.

One point of context first. The source paper defines its *inactivated* state on
*PDGFRB*, *TAGLN*, *ACTA2* and *PPARG*, so *ACTA2* and *TAGLN* are markers of
that state rather than of activation. Loss of those genes from the stellate
compartment therefore has two candidate explanations, mural contamination or
genuine loss of an inactivated stellate population, and they are separable
because they predict different behavior from smooth-muscle genes that appear in
neither panel.

```{r}
#| label: stellate-subtypes

sub_expr <- cached("stellate_subtype_expr.rds", {
  library(BPCells)
  mat  <- open_matrix_dir("bpcells_counts")
  meta <- readRDS(cache_path("meta.rds"))
  sel  <- grepl("^Stellate", meta[[COL_CT_FINE]])
  ctf  <- factor(as.character(meta[[COL_CT_FINE]])[sel])
  sb   <- mat[, which(sel)]
  nm   <- log1p(multiply_cols(sb, 1e4 / colSums(sb)))
  g <- c("DCN", "RBP1", "LRAT", "CYGB", "HGF", "COL1A1", "COL1A2",
         "PDGFRB", "TAGLN", "ACTA2", "PPARG",
         "MYH11", "MYLK", "MYL9", "NOTCH3", "RGS5")
  g <- g[g %in% rownames(mat)]
  t(pseudobulk_matrix(nm[g, ], cell_groups = ctf, method = "mean"))
})

rownames(sub_expr) <- sub("Stellate_cell_", "", rownames(sub_expr))
kable(round(sub_expr[, c("LRAT", "DCN", "RBP1", "COL1A1",
                         "ACTA2", "TAGLN", "PDGFRB",
                         "MYH11", "NOTCH3", "RGS5")], 2),
      caption = "Mean log-normalized expression by stellate subcluster. Left block: hepatic stellate identity. Middle: the paper's inactivated-state panel. Right: smooth-muscle and pericyte genes used by neither panel.")
```

The inactivated subcluster carries
`r round(sub_expr["inactivated", "MYH11"] / max(sub_expr[rownames(sub_expr) != "inactivated", "MYH11"]))`-fold
more *MYH11* than any other stellate subcluster, along with the only appreciable
*NOTCH3*, while showing the **lowest** *LRAT* and *DCN* of the four.

That combination decides the question. *MYH11* is smooth-muscle myosin heavy
chain and is not expressed by hepatic stellate cells in any activation state. A
genuinely inactivated stellate cell, one that has reverted after activation,
retains its lipid-storing identity markers; this population has the least *LRAT*
and *DCN* of any subcluster while uniquely expressing *MYH11* and *NOTCH3*. It
is better described as vascular smooth muscle or pericyte than as a stellate
cell of any state.

That mural population also declines across the progression, falling from
`r round(mean(st_frac$pct[st_frac$subtype == "Stellate_cell_inactivated" & st_frac$stage == "No-MASLD"], na.rm = TRUE), 1)`%
to
`r round(mean(st_frac$pct[st_frac$subtype == "Stellate_cell_inactivated" & st_frac$stage == "aMASH"], na.rm = TRUE), 1)`%
of the compartment, which is the right shape to inflate an apparent loss of
contractile genes when the whole label is aggregated.

### How much of the effect is composition?

The tempting conclusion is that the entire contractile signal is a mixture
artifact. That is testable in two ways, and the test does not fully support it.

```{r}
#| label: stellate-decompose

# 1. Rebuild the stellate pseudobulk with the mural subcluster removed.
# 2. Test inside single subtypes, where composition is fixed by construction and
#    any remaining change must be a within-cell-type effect.
decomp <- cached("stellate_decompose.rds", {
  library(BPCells); library(DESeq2)
  meta <- readRDS(cache_path("meta.rds")); mt <- open_matrix_dir("bpcells_counts")
  dn <- unique(meta[, c(COL_DONOR, COL_STAGE)])
  cnd <- setNames(unname(CONDITION_MAP[dn[[COL_STAGE]]]), dn[[COL_DONOR]])

  de_on <- function(subtypes) {
    sel <- meta[[COL_CT_FINE]] %in% subtypes
    grp <- paste0("g__", meta[[COL_DONOR]][sel])
    p <- pseudobulk_matrix(mt[, which(sel)], cell_groups = grp, method = "sum")
    n <- table(grp)[colnames(p)]; dono <- sub("g__", "", colnames(p))
    keep <- n >= MIN_CELLS & !is.na(cnd[dono])
    cts <- round(p[, keep, drop = FALSE]); mode(cts) <- "integer"
    cts <- cts[rowSums(cts > 0) >= ceiling(ncol(cts) / 3), , drop = FALSE]
    cd <- data.frame(condition = factor(cnd[dono[keep]], levels = CONDITION_LEVELS),
                     row.names = colnames(cts))
    dds <- DESeq(DESeqDataSetFromMatrix(cts, cd, ~ condition), quiet = TRUE)
    r <- as.data.frame(results(dds, contrast = c("condition", "MASH", "Control")))
    r$gene <- rownames(r); r$n_control <- sum(cd$condition == "Control")
    r$n_mash <- sum(cd$condition == "MASH"); r
  }
  list(
    hsc_only  = de_on(c("Stellate_cell_quiescent", "Stellate_cell_activated",
                        "Stellate_cell_misc")),
    quiescent = de_on("Stellate_cell_quiescent"),
    activated = de_on("Stellate_cell_activated"))
})

pull <- function(r, g) {
  i <- match(g, r$gene)
  sprintf("%+.2f%s", r$log2FoldChange[i],
          ifelse(is_de(r)[i], "*", ""))
}
gg <- c("TAGLN", "ACTA2", "MYH11", "MYL9", "LRAT", "COL1A1", "COL3A1")
kable(data.frame(
  Gene = gg,
  `All stellate` = pull(de$Stellate_cell, gg),
  `Mural removed` = pull(decomp$hsc_only, gg),
  `Quiescent only` = pull(decomp$quiescent, gg),
  `Activated only` = pull(decomp$activated, gg),
  check.names = FALSE), row.names = FALSE,
  caption = paste0("log2 fold change, MASH vs Control, as the mural subcluster is ",
                   "removed and then as the test is confined within single subtypes. * = ", DE_RULE, "."))
```

The two contractile genes behave differently, and the distinction matters.

*MYH11* loses significance as soon as the mural subcluster is removed and stays
non-significant inside both subtypes. That is what a pure composition effect
looks like, and it confirms the mural reading.

*TAGLN* does not. It remains strongly depleted inside the quiescent subcluster
and inside the activated subcluster, tested separately, where composition is
fixed by construction and cannot contribute. That is a real transcriptional
change within hepatic stellate cells.

So both mechanisms are operating, and an earlier reading of this result as
purely a mixture artifact would have been wrong. A mural population is inflating
the aggregate signal, which is why `MYOGENESIS` weakens from
`r round(subset(gs_all, cell_type == "Stellate_cell" & pathway == "MYOGENESIS")$NES, 2)`
to non-significance once those cells are excluded. Underneath it there is a
genuine, subtype-independent downregulation of *TAGLN* in stellate cells. What
the data do not support is the simple statement that stellate cells are losing
their activation program wholesale: *LRAT* rises rather than falls, and the
fibrillar collagens do not move significantly in either direction. That last
point should not be read as evidence against fibrogenesis either: as quantified
in the limitations, this dataset contains no collagen-producing myofibroblast
population to detect a change in.

This remains a caution about the deposited annotation, and it is only visible
because the deposit ships substate labels alongside the broad ones. It is also a
concrete illustration of the general hazard, with the qualification that
decomposing the effect matters: pseudobulk over a mixed label can exaggerate a
program, and separating composition from within-cell-type change requires
testing at a level where composition cannot vary.

## Comparison with the source paper's own enrichment

The source paper ran its own enrichment on the same tissue, reported in its
Supplementary Table 3. Comparing against it tests whether the pathway-level
conclusions here depend on the analytical choices made in this report.

Four differences matter when reading the comparison. Their differential
expression was cell-level Seurat `FindMarkers`, against donor-level pseudobulk
here. Their enrichment was over-representation testing with enrichR on
thresholded gene lists, against GSEA on the full ranking here. Their gene sets
were GO Biological Process, against Hallmark here, and the two collections do
not map one-to-one. And their comparison pooled MASL into the reference group as
"nonMASH", where the primary contrast here holds MASL out. Their table also
reports only the top four terms in each direction, so a term missing from it is
not evidence that they found nothing.

```{r}
#| label: paper-comparison

# Terms as reported in Hong et al. Supplementary Table 3 (GO Biological Process,
# enrichR, MASH vs nonMASH). Quoted here as reference values; every value in the
# "this report" columns is computed from the GSEA above.
paper <- data.frame(
  cell_type = c("Hepatocyte", "Hepatocyte", "Hepatocyte", "Hepatocyte",
                "Cholangiocyte", "Cholangiocyte", "Cholangiocyte",
                "Stellate_cell", "Stellate_cell",
                "Endothelial_cell", "Endothelial_cell"),
  paper_term = c("cholesterol biosynthetic process", "steroid metabolic process",
                 "drug catabolic process", "cellular amino acid catabolic process",
                 "mitochondrial electron transport", "Notch signaling pathway",
                 "regulation of canonical Wnt signaling",
                 "muscle contraction", "collagen fibril organization",
                 "cellular response to interferon-gamma",
                 "positive regulation of cell migration"),
  paper_dir = c("up", "down", "down", "down",
                "down", "up", "up",
                "down", "up",
                "down", "up"),
  hallmark = c("CHOLESTEROL_HOMEOSTASIS", "BILE_ACID_METABOLISM",
               "XENOBIOTIC_METABOLISM", "FATTY_ACID_METABOLISM",
               "OXIDATIVE_PHOSPHORYLATION", "NOTCH_SIGNALING",
               "WNT_BETA_CATENIN_SIGNALING",
               "MYOGENESIS", "EPITHELIAL_MESENCHYMAL_TRANSITION",
               "INTERFERON_GAMMA_RESPONSE", "EPITHELIAL_MESENCHYMAL_TRANSITION"),
  stringsAsFactors = FALSE)

lookup <- function(ct, pw, field) {
  r <- gs_all[gs_all$cell_type == ct & gs_all$pathway == pw, ]
  if (!nrow(r)) return(NA_real_)
  r[[field]][1]
}
paper$NES <- mapply(lookup, paper$cell_type, paper$hallmark, "NES")
paper$FDR <- mapply(lookup, paper$cell_type, paper$hallmark, "padj")
paper$this_dir <- ifelse(is.na(paper$NES), NA,
                  ifelse(paper$FDR >= 0.05, "n.s.",
                  ifelse(paper$NES > 0, "up", "down")))
paper$agrees <- ifelse(paper$this_dir == "n.s.", "not significant",
                ifelse(paper$this_dir == paper$paper_dir, "agrees", "opposite"))

kable(data.frame(
  `Cell type` = paper$cell_type,
  `Paper GO term` = paper$paper_term,
  `Paper` = paper$paper_dir,
  `Matched Hallmark set` = paper$hallmark,
  `NES` = round(paper$NES, 2),
  `FDR` = signif(paper$FDR, 2),
  `Verdict` = paper$agrees,
  check.names = FALSE), row.names = FALSE,
  caption = "Terms reported in Hong et al. Supplementary Table 3, against the nearest Hallmark set in this analysis.")
```

**Hepatocyte metabolic reprogramming replicates completely.** All four reported
directions agree: cholesterol handling up, and steroid, drug-catabolic and
fatty-acid metabolism down. Two methods that share nothing except the underlying
tissue reach the same conclusion, which is the strongest form of corroboration
available here.

**Their table contains the stellate result independently.** Supplementary Table 3
reports *muscle contraction* as significantly depleted in stellate cells in
MASH. That is the same signal as the `MYOGENESIS` depletion above
(NES `r round(subset(gs_all, cell_type == "Stellate_cell" & pathway == "MYOGENESIS")$NES, 2)`),
recovered from different data processing, a different differential expression
method and a different gene set collection.

This matters for how the mural-cell interpretation should be read. The
observation is not particular to this pipeline: it is present in the source
paper's own supplementary results, where it is reported without interpretation.
What this report adds is the explanation, that the depletion reflects loss of a
mural population carried inside the stellate label rather than a change in
stellate contractile state.

### Where the two analyses diverge

```{r}
#| label: paper-grouping

# Their reference group pools MASL with controls. Rerunning the four cell types
# their table covers under that definition separates disagreements caused by the
# grouping from disagreements caused by method.
gsea_pooled <- cached("gsea_pooled.rds", {
  library(fgsea); library(msigdbr)
  hall <- msigdbr(species = "Homo sapiens", collection = "H")
  sets <- split(hall$gene_symbol, hall$gs_name)
  names(sets) <- sub("^HALLMARK_", "", names(sets))
  pm <- pbmeta
  pm$grp <- ifelse(pm$stage %in% c("No-MASLD", "MASL"), "nonMASH",
            ifelse(pm$stage %in% c("eMASH", "aMASH"), "MASH", NA))
  out <- list()
  for (ct in c("Hepatocyte", "Cholangiocyte", "Stellate_cell", "Endothelial_cell")) {
    sel <- pm$cell_type == ct & pm$n_cells >= MIN_CELLS & !is.na(pm$grp)
    m <- pm[sel, ]; m$grp <- factor(m$grp, levels = c("nonMASH", "MASH"))
    cts <- round(pb[, m$pseudosample, drop = FALSE]); mode(cts) <- "integer"
    cts <- cts[rowSums(cts > 0) >= ceiling(ncol(cts) / 3), , drop = FALSE]
    dds <- DESeq(DESeqDataSetFromMatrix(cts, m, ~ grp), quiet = TRUE)
    r <- as.data.frame(results(dds, contrast = c("grp", "MASH", "nonMASH")))
    r$gene <- rownames(r); r <- r[!is.na(r$stat), ]
    st <- setNames(r$stat, r$gene); st <- st[!duplicated(names(st))]
    set.seed(42)
    out[[ct]] <- as.data.frame(fgsea(sets, st, minSize = 10, maxSize = 500,
                                     nPermSimple = 10000))
  }
  out
})

pooled_get <- function(ct, pw) {
  r <- gsea_pooled[[ct]][gsea_pooled[[ct]]$pathway == pw, ]
  if (!nrow(r)) return(c(NA, NA))
  c(r$NES[1], r$padj[1])
}
diverge <- data.frame(
  `Cell type` = c("Stellate_cell", "Endothelial_cell", "Cholangiocyte", "Cholangiocyte"),
  `Hallmark set` = c("EPITHELIAL_MESENCHYMAL_TRANSITION", "INTERFERON_GAMMA_RESPONSE",
                     "NOTCH_SIGNALING", "WNT_BETA_CATENIN_SIGNALING"),
  check.names = FALSE)
mine <- t(mapply(function(ct, pw) c(lookup(ct, pw, "NES"), lookup(ct, pw, "padj")),
                 diverge$`Cell type`, diverge$`Hallmark set`))
pool <- t(mapply(pooled_get, diverge$`Cell type`, diverge$`Hallmark set`))
diverge$`NES (MASL held out)`   <- round(mine[, 1], 2)
diverge$`FDR (held out)`        <- signif(mine[, 2], 2)
diverge$`NES (MASL pooled in)`  <- round(pool[, 1], 2)
diverge$`FDR (pooled in)`       <- signif(pool[, 2], 2)
kable(diverge, row.names = FALSE,
      caption = "The four disagreements, evaluated under both reference-group definitions.")
```

**One disagreement is entirely explained by the grouping.** The paper reports
collagen fibril organization up in stellate cells. Holding MASL out, the nearest
Hallmark set is not significant; pooling MASL into the reference as they did, it
becomes significantly up. The reason is visible in the composition data above:
MASL donors carry an unusually quiescent stellate compartment
(`r round(mean(st_frac$pct[st_frac$subtype == "Stellate_cell_quiescent" & st_frac$stage == "MASL"], na.rm = TRUE))`%
quiescent, against
`r round(mean(st_frac$pct[st_frac$subtype == "Stellate_cell_quiescent" & st_frac$stage == "No-MASLD"], na.rm = TRUE))`%
in controls), so folding them into the reference lowers its fibrogenic baseline
and widens the contrast. Both results are defensible; they answer slightly
different questions, and the fibrogenesis signal is worth knowing is sensitive
to that choice.

**One disagreement is not.** The paper reports the interferon-gamma response
down in endothelial cells; here it is up, under both reference definitions. The
leading edge is *VCAM1*, *CD274*, *IDO1*, *FAS* and *IL18BP*, which are
interferon-inducible endothelial activation genes, and *VCAM1* induction is
consistent with the leukocyte recruitment implied by the abundance results. The
direction is also coherent across the tissue: interferon-gamma response is up in
the structural compartments and down in every immune population tested. The most
likely source of the discrepancy is that Hallmark's response signature is broad
where the GO term is a narrower signalling set, compounded by cell-level versus
donor-level differential expression. This is flagged rather than resolved.

**Two of their findings are not reproduced.** Notch and Wnt signalling in
cholangiocytes are not significant here under either grouping. Cholangiocytes
are the least powered cell type in this analysis, with the fewest donors
clearing the pseudo-sample threshold, so this is weak evidence against their
result. Their Notch claim also rests primarily on cell-cell interaction
analysis, which is not attempted here.


```{r}
#| label: session
#| code-fold: true
sessionInfo()
```
